---
title: "A Right That Exists on Paper"
subtitle: "Compassionate release, and the distance between a law and its use"
format:
  html:
    toc: true
    toc-depth: 2
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(scales)

# Every figure on this page comes from a public document: federal sentencing
# reports, published state-policy research, and state statutes. Each CSV row
# carries its own source and the specific table or page it came from, and the
# guard below refuses to render if any row is missing one. No case file, no
# person-level record, and no unpublished data feeds anything here.
fed_month  <- read.csv("data/federal_motions_monthly.csv", fileEncoding = "UTF-8")
fed_reason <- read.csv("data/federal_reasons.csv", fileEncoding = "UTF-8")
pol_types  <- read.csv("data/state_policy_types.csv", fileEncoding = "UTF-8")
reqs       <- read.csv("data/state_requirements.csv", fileEncoding = "UTF-8")
util       <- read.csv("data/state_utilization.csv", fileEncoding = "UTF-8")
ages       <- read.csv("data/state_age_thresholds.csv", fileEncoding = "UTF-8")
wi         <- read.csv("data/wisconsin_timeline.csv", fileEncoding = "UTF-8")

fed_month <- fed_month |>
  mutate(
    date  = as.Date(sprintf("%d-%02d-01", year, month_num)),
    label = paste0(substr(month, 1, 3), " ", year)
  ) |>
  arrange(date)

# The reason codes are the Sentencing Commission's own. The theme grouping on
# top of them is mine, and the codebook section on this page spells out every
# assignment so a reader can disagree with a specific one.
reason_theme <- fed_reason |>
  group_by(outcome, theme) |>
  summarise(n = sum(n), .groups = "drop") |>
  group_by(outcome) |>
  mutate(share = n / sum(n)) |>
  ungroup()

funnel   <- util |> filter(chart == "funnel")   |> arrange(stage)
tracking <- util |> filter(chart == "tracking") |> arrange(stage)
notice   <- util |> filter(chart == "notice")

age_span  <- ages |> filter(measure == "minimum_age")
age_count <- ages |> filter(measure == "count")

# The age-span chart draws directly from the CSV rows, so a re-coded state
# layer changes the chart without touching this file.
span_df <- age_span |>
  mutate(y = 1, lab = paste0(state, "\n", value))

pol_types <- pol_types |>
  mutate(policy_type_f = factor(policy_type, levels = rev(policy_type)))

# Wrap the longest requirement names so the icon array keeps its width for
# the dots rather than the axis labels.
wrap_lab <- function(s, width) {
  vapply(s, function(x) paste(strwrap(x, width = width), collapse = "\n"),
         character(1), USE.NAMES = FALSE)
}

reqs <- reqs |>
  mutate(requirement_f = factor(requirement, levels = rev(requirement),
                                labels = wrap_lab(rev(requirement), 16)))

# One row per state per requirement: 46 units, filled while unit <= n. The
# icon array reads from these.
req_units <- reqs |>
  uncount(46, .id = "unit") |>
  mutate(met = unit <= n)

wi <- wi |> arrange(sort_year)

# The tooltips quote source_detail from every file, so a blank one would
# render as an empty citation.
detail_all <- c(fed_month$source_detail, fed_reason$source_detail,
                pol_types$source_detail, reqs$source_detail,
                util$source_detail, ages$source_detail, wi$source_detail)

# Fail the CI render loudly rather than publish a wrong page (a failed publish
# leaves the previous version of the site live). The totals below are the ones
# the source documents state for themselves, so a transcription slip in any
# single row breaks the sum and stops the build.
stopifnot(
  nrow(fed_month) == 54,
  !anyNA(fed_month$date), !anyNA(fed_month$total),
  sum(fed_month$total) == 15265, sum(fed_month$granted) == 2088,
  all(fed_month$granted + fed_month$denied == fed_month$total),
  all(fed_month$granted_pct >= 0 & fed_month$granted_pct <= 100),

  nrow(fed_reason) == 45,
  all(fed_reason$outcome %in% c("Granted", "Denied")),
  sum(fed_reason$n[fed_reason$outcome == "Granted"]) == 631,
  sum(fed_reason$n[fed_reason$outcome == "Denied"]) == 5067,
  all(fed_reason$n > 0),
  nrow(reason_theme) == 13,
  all(abs(tapply(reason_theme$share, reason_theme$outcome, sum) - 1) < 1e-9),

  nrow(pol_types) == 7, !anyNA(pol_types$policy_type_f),
  nrow(reqs) == 8, !anyNA(reqs$requirement_f),
  all(reqs$n <= 46), all(reqs$pct > 0 & reqs$pct <= 100),
  nrow(req_units) == 8 * 46, sum(req_units$met) == sum(reqs$n),

  nrow(funnel) == 3, nrow(tracking) == 3,
  funnel$value[1] == 5932, funnel$value[2] == 2751, funnel$value[3] == 802,
  all(diff(funnel$value) < 0),
  all(tracking$value <= 49),
  nrow(notice) == 1, notice$value[1] == 2,

  nrow(age_span) == 2, nrow(age_count) == 2,
  min(age_span$value) == 55, max(age_span$value) == 70,

  nrow(wi) == 8, !anyNA(wi$sort_year), all(diff(wi$sort_year) > 0),

  # Every row on this page must be able to say where it came from.
  !any(fed_month$source == ""), !any(fed_reason$source == ""),
  !any(pol_types$source == ""), !any(reqs$source == ""),
  !any(util$source == ""), !any(ages$source == ""), !any(wi$source == ""),
  !anyNA(detail_all), !any(detail_all == "")
)

plum_dark   <- "#6b4f6c"
plum        <- "#8C6A8D"
plum_series <- "#774379"   # chroma-bumped plum: only for multi-series charts
coral_dark  <- "#c96b52"
gray_ctx    <- "#8a8a8a"   # de-emphasized context; always direct-labeled

theme_cr <- function() {
  theme_minimal(base_family = "sans", base_size = 12) +
    theme(
      plot.title    = element_text(size = 14, face = "bold", color = "#3C3C3C"),
      plot.subtitle = element_text(size = 11, color = "#555555"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#eeeeee"),
      axis.text     = element_text(color = "#444444"),
      legend.position = "bottom",
      legend.title  = element_text(size = 10, face = "bold"),
      strip.text    = element_text(face = "bold", color = "#333333")
    )
}

# ggplotly() drops ggplot subtitles. Fold each subtitle into the plotly title
# as a smaller <br><sup> second line so it survives the conversion. Pipe every
# chart through this instead of bare ggplotly().
# The tail fixes a sizing failure: plotly falls back to a fixed 700px when its
# container has no width at draw time (a background tab, a layout still
# settling), and nothing re-draws until a window resize, so the chart sits
# cropped inside the card's overflow: hidden. config(responsive) re-draws on
# window resizes; the ResizeObserver re-draws when the container gets its
# real width without one.
ggplotly_titled <- function(p, ...) {
  fig <- ggplotly(p, ...)
  sub <- p$labels$subtitle
  if (!is.null(sub) && nzchar(sub)) {
    fig <- fig |>
      layout(
        title = list(text = paste0(
          "<b>", p$labels$title, "</b><br><sup>", sub, "</sup>"
        )),
        margin = list(t = 70)
      )
  }
  fig |>
    config(responsive = TRUE) |>
    htmlwidgets::onRender("
      function(el) {
        if (!window.ResizeObserver) return;
        new ResizeObserver(function() {
          var w = el.parentElement.getBoundingClientRect().width;
          if (w > 0 && el._fullLayout &&
              Math.abs(w - el._fullLayout.width) > 1) {
            Plotly.Plots.resize(el);
          }
        }).observe(el.parentElement);
      }
    ")
}

clear_bg <- function(fig) {
  layout(fig, paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
}

# ggplotly parks a horizontal legend on top of the x-axis title. Anchor it
# below the title and pad the bottom margin. Pipe every chart that has a
# legend through this; legend-free charts do not need it. The y offset is
# relative to plot height, so shorter charts need a deeper y (and more b)
# to clear the title by the same number of pixels.
legend_below <- function(fig, y = -0.28, b = 95) {
  layout(fig,
         legend = list(orientation = "h", x = 0, y = y, yanchor = "top"),
         margin = list(b = b))
}

# ggplotly drops a constant hjust on geom_text: textposition comes back null
# and plotly centers each label on its anchor, which runs labels into bars,
# dots, and each other. Re-anchor every text trace after conversion; the sign
# of the anchor x picks the side. Pipe any chart whose geom_text sets hjust
# through this; charts with centered labels stay off it.
anchor_text <- function(fig) {
  fig$x$data <- lapply(fig$x$data, function(tr) {
    if (!is.null(tr$mode) && grepl("text", tr$mode, fixed = TRUE)) {
      tr$textposition <- if (tr$x[[1]] > 0) "middle right" else "middle left"
    }
    tr
  })
  fig
}
```

```{=html}
<style>
/* This page runs chart-forward: section prose steps down a size, and the
   card ledes (styled in site-theme.scss) carry each section's setup. */
section.level2 > p, main.content > p { font-size: 0.95rem; }
/* The codebook table is wider than a phone. Let it scroll inside its own
   box so the page body never does. */
.table-scroll { overflow-x: auto; }
.table-scroll table { font-size: 0.86rem; }
</style>
```

Almost every state has a law for releasing people from prison when they
are dying, seriously ill, or old enough that the risk they once posed has
passed. The names and tests vary, but the premise is shared: a sentence
can outlive its purpose, and the law should have a way to say so.

Almost none of these laws see much use. This page measures the distance
between the right on paper and the right in practice, and it does so
entirely from public records.

It is also a working demonstration of a method. Statutes and case
decisions are qualitative data, and the usual way to present them is a
memo. But most questions people bring to a law are comparative: how many
states require a diagnosis, how far apart are the age cutoffs, what
reasons do courts actually give, what changed and when. Those are chart
questions, so each section asks one, answers it with a chart form suited
to categorical and legal evidence, and notes why that form got the job.

## Data Sources {.section-heading}

Everything here comes from documents that are public, and part of the
point is to show how far public records can carry a question like this
one. Each data file behind the charts names its source on every row, and
the build refuses to render a row that does not.

**Federal decisions.** U.S. Sentencing Commission, [Compassionate Release
Data Reports](https://www.ussc.gov/research/data-reports/compassionate-release-data-reports):
the [Fiscal Year 2025 report](https://www.ussc.gov/sites/default/files/pdf/research-and-publications/federal-sentencing-statistics/compassionate-release/FY25-Compassionate-Release.pdf)
(tables 1, 10, and 11) and the [2nd Quarter FY2026 report](https://www.ussc.gov/sites/default/files/pdf/research-and-publications/federal-sentencing-statistics/compassionate-release/FY26Q2-Compassionate-Release.pdf)
(table 1). These are court decisions under 18 U.S.C. sec. 3582(c)(1)(A),
which since the First Step Act of 2018 may be sought by a sentenced person
directly; they feed the monthly docket chart and the reasons analysis.
Closed fiscal years follow the full-year report, and fiscal year 2026 data
is preliminary and cumulative through the second quarter.

**State policy content.** Holland, Prost, Hoffmann, and Dickinson (2020),
["U.S. Department of Corrections Compassionate Release Policies: A Content
Analysis and Call to Action"](https://doi.org/10.1177/0030222818791708),
<em>OMEGA: Journal of Death and Dying</em> 81(4), 607 to 626. Their coding
covers all fifty states from policies collected in 2016, which makes it
both the most complete state-level source available and, by now, a decade
old. It feeds the policy-type, requirement, and age charts; counts are
cited to it, and the figures are drawn fresh rather than reproduced.

**State utilization.** Holland, Prost, Hoffmann, and Dickinson (2021),
["Access and Utilization of Compassionate Release in State Departments of
Corrections"](https://doi.org/10.1080/13576275.2020.1750357),
<em>Mortality</em> 26(1), 49 to 65, built from public records requests
covering 2013 to 2015. It feeds the eligible-to-released funnel and the
record-keeping tiles.

**Wisconsin.** The instruments themselves: [Wis. Stat. sec.
302.113(9g)](https://docs.legis.wisconsin.gov/statutes/statutes/302/113/9g),
[Wis. Admin. Code DOC sec. 302.41](https://docs.legis.wisconsin.gov/document/administrativecode/DOC%20302.41),
court form [CR-254](https://www.wicourts.gov/forms/CR-254.pdf) and its
companion orders, and [DAI Policy
302.00.13](https://doc.wi.gov/DepartmentPoliciesDAI/3020013.pdf). They
feed the one-state timeline, and they are a tidy demonstration that "the
law" on any given point is several documents, kept by different offices.

## Nearly Everywhere on Paper {.section-heading}

::: {.chart-card}
<p class="chart-card-lede">Start with what exists on paper. When
researchers last read all fifty states' policies in one pass, in 2016,
what kinds of release mechanism did they find, and how many states had
none?</p>
```{r chart-policy-types}
#| fig-height: 3.8
#| out-width: "100%"
p_types <- ggplot(pol_types, aes(x = n, y = policy_type_f)) +
  geom_segment(aes(x = 0, xend = n, yend = policy_type_f),
               color = "#dcd5dc", linewidth = 0.9) +
  geom_point(aes(text = paste0(
    "<b>", policy_type, "</b><br>",
    n, " states<br>",
    "Source: ", source, "<br>Policies as of ", data_year
  )),
  size = 4,
  color = ifelse(pol_types$policy_type == "No mechanism", coral_dark, plum_dark)) +
  geom_text(aes(label = n), nudge_x = 1.6, size = 3.4, color = "#3C3C3C") +
  scale_x_continuous(limits = c(0, 36)) +
  labs(
    title = "Four states had nothing; the rest had something",
    subtitle = "Types of release mechanism identified across the fifty states, policies as of 2016",
    x = "Number of states", y = NULL
  ) +
  theme_cr()

ggplotly_titled(p_types, tooltip = "text") |> clear_bg()
```
<p class="chart-card-note">Forty-six states had a compassionate release
mechanism; four did not (Illinois, Iowa, Nebraska, and Utah). Counts sum
to more than fifty because a state can hold more than one kind of policy,
so read these as counts of policy types, not as the fifty states sorted
into piles.
Source: Holland, Prost, Hoffmann, and Dickinson (2020), <em>OMEGA</em>
81(4), figure 1 at p. 613.</p>
<p class="chart-card-note">The picture has since shifted. Illinois passed
the Joe Coleman Medical Release Act in 2021, effective at the start of
2022. Nebraska, where medical parole for terminal illness or permanent
incapacity has existed since 2005, added geriatric parole in 2023. That
is the trouble with a decade-old map: it is the best one that exists, and
it is already wrong.</p>
<p class="chart-card-note"><strong>Why this chart:</strong> a lollipop
rather than a bar, because the quantity is a count of states and the bar
area would add ink without adding information. The one category that is an
absence rather than a policy is the only one in coral, so the exception is
visible before the labels are read.</p>
:::

## What You Have to Prove {.section-heading}

::: {.chart-card}
<p class="chart-card-lede">Having a policy is one thing; qualifying under
it is another. Across the forty-six states with a policy, what did the
policies require?</p>
```{r chart-requirements}
#| fig-height: 3.9
#| out-width: "100%"
p_req <- ggplot(req_units, aes(x = unit, y = requirement_f)) +
  geom_point(aes(color = met, text = paste0(
    "<b>", requirement, " requirement</b><br>",
    n, " of 46 states (", pct, "%)<br>",
    "Source: ", source
  )), size = 2) +
  geom_text(data = reqs, aes(x = 48, y = requirement_f, label = n),
            inherit.aes = FALSE, hjust = 0, size = 3.3, color = "#3C3C3C") +
  scale_color_manual(values = c(`TRUE` = plum_dark, `FALSE` = "#e2dbe2"),
                     guide = "none") +
  scale_x_continuous(limits = c(0, 54), breaks = NULL) +
  labs(
    title = "Nearly all required a diagnosis; two thirds required a risk finding",
    subtitle = "Each dot is one of the 46 states with a policy; filled dots imposed the requirement, as of 2016",
    x = NULL, y = NULL
  ) +
  theme_cr() +
  theme(panel.grid.major = element_blank())

ggplotly_titled(p_req, tooltip = "text") |> clear_bg() |> anchor_text()
```
<p class="chart-card-note">Source: Holland, Prost, Hoffmann, and Dickinson
(2020), <em>OMEGA</em> 81(4), table 1 at p. 612. One caution carried from
the source: the table reports age restrictions in nine states and the
article text reports ten. The chart follows the table and flags the
difference rather than quietly picking a side.</p>
<p class="chart-card-note"><strong>Why this chart:</strong> an icon array,
the form risk communication research keeps reaching for when a share has
to stay countable. Each dot is a state, so 45 of 46 reads as one state shy
of everyone rather than as a long bar, and the denominator never leaves
the picture. Sorting by prevalence rather than alphabetically turns a list
of requirements into a finding.</p>
:::

## How Old Is Old Enough {.section-heading}

::: {.chart-card}
<p class="chart-card-lede">Where age enters these laws, it enters as a
specific number, and states did not pick the same one.</p>
```{r chart-age-span}
#| fig-height: 2.4
#| out-width: "100%"
p_age <- ggplot(span_df, aes(x = value, y = y)) +
  geom_segment(x = min(span_df$value), xend = max(span_df$value),
               y = 1, yend = 1,
               color = "#dcd5dc", linewidth = 1.6) +
  geom_point(aes(text = paste0(
    "<b>", state, "</b><br>Minimum age ", value,
    "<br>Source: Holland et al. (2020), ", source_detail
  )), color = plum_dark, size = 5) +
  geom_text(aes(label = lab), nudge_y = 0.42, size = 3.4,
            color = "#3C3C3C", lineheight = 0.95) +
  scale_x_continuous(limits = c(50, 75), breaks = seq(50, 75, 5)) +
  scale_y_continuous(limits = c(0.55, 1.75)) +
  labs(
    title = "Fifteen years separate the youngest cutoff from the oldest",
    subtitle = "Minimum qualifying age among state policies that set one, as of 2016",
    x = "Age", y = NULL
  ) +
  theme_cr() +
  theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank())

ggplotly_titled(p_age, tooltip = "text") |> clear_bg()
```
<p class="chart-card-note">Ten states used age as a factor at all, and in
only three could age decide the question on its own. Wyoming's policy set
no number, requiring instead that a person be "incapacitated by age."
Source: Holland, Prost, Hoffmann, and Dickinson (2020), <em>OMEGA</em>
81(4), p. 614.</p>
<p class="chart-card-note">The same boundary problem shows up whenever a
law puts a number on a fuzzy category, which is the subject of a short
exercise on this site:
<a href="../howold/how_old_is_old.html">How Old Is Old?</a></p>
<p class="chart-card-note"><strong>Why this chart:</strong> a span rather
than a bar chart of two values, because the finding is the distance
between the endpoints, so the distance is what gets drawn. Two bars would
have asked the reader to compare heights, and the height comparison is not
the question.</p>
:::

## Eligible, Applied, Released {.section-heading}

::: {.chart-card}
<p class="chart-card-lede">From the law to its use: of everyone the states
themselves counted as eligible over three years, how many applied, and how
many were released?</p>
```{r chart-funnel}
#| fig-height: 3
#| out-width: "100%"
fun <- funnel |>
  mutate(
    category_f = factor(category, levels = rev(category)),
    pct_of_elig = value / max(value),
    # The eligible bar is its own denominator, so its label skips the trivial
    # 100.0% and leaves room for the label at phone widths.
    lab = ifelse(stage == 1, comma(value),
                 paste0(comma(value), "  (",
                        percent(pct_of_elig, accuracy = 0.1), ")"))
  )

p_fun <- ggplot(fun, aes(x = value, y = category_f)) +
  geom_col(aes(text = paste0(
    "<b>", category, "</b><br>", comma(value), " people<br>",
    percent(pct_of_elig, accuracy = 0.1), " of those eligible<br>",
    "Source: ", source
  )), fill = c("#e2dbe2", plum, plum_dark), width = 0.62) +
  geom_text(aes(label = lab), hjust = 0, nudge_x = 90,
            size = 3.4, color = "#3C3C3C") +
  scale_x_continuous(limits = c(0, 7600), labels = comma) +
  labs(
    title = "Of everyone eligible, one in seven got out",
    subtitle = "State compassionate release, 2013 to 2015, among states that reported each stage",
    x = "People", y = NULL
  ) +
  theme_cr()

ggplotly_titled(p_fun, tooltip = "text") |> clear_bg() |> anchor_text()
```
<p class="chart-card-note">Source: Holland, Prost, Hoffmann, and Dickinson
(2021), <em>Mortality</em> 26(1), pp. 55 to 56. Denominators are reporting
states only, so this is the funnel among systems that could produce the
number at all. Note also what the last bar is not: the share who were
refused. People leave this process by dying before a decision and by
withdrawing, so the gap between applying and release is wider than any
denial rate.</p>
<p class="chart-card-note"><strong>Why this chart:</strong> a funnel drawn
as plain nested bars on a shared axis, because each stage is a subset of
the one above it and the reader should be able to measure the drop, not
just sense it. Tapered funnel graphics encode the same numbers as areas
that no one can compare. The shading deepens as the pool narrows, the same
part-in-whole grammar the federal chart below uses.</p>
:::

## Does Anyone Have to Count? {.section-heading}

::: {.chart-card}
<p class="chart-card-lede">Every number in the section above depends on a
state having written something down. Most of the work of finding out
whether a law functions is finding out whether anyone kept the record.</p>
<div style="display:flex; flex-wrap:wrap; gap:0.75rem; margin:0.4rem 0 1rem;">
<div style="flex:1 1 150px; min-width:150px; text-align:center; padding:0.95rem 0.5rem; background:#FDFBF7; border-radius:8px;">
<div style="font-size:1.8rem; font-weight:700; color:#2E1A47;">`r tracking$value[1]` of 49</div>
<div style="font-size:0.8rem; color:#666;">tracked how many people they released</div>
</div>
<div style="flex:1 1 150px; min-width:150px; text-align:center; padding:0.95rem 0.5rem; background:#FDFBF7; border-radius:8px;">
<div style="font-size:1.8rem; font-weight:700; color:#2E1A47;">`r tracking$value[2]` of 49</div>
<div style="font-size:0.8rem; color:#666;">tracked how many people applied</div>
</div>
<div style="flex:1 1 150px; min-width:150px; text-align:center; padding:0.95rem 0.5rem; background:#FDFBF7; border-radius:8px;">
<div style="font-size:1.8rem; font-weight:700; color:#2E1A47;">`r tracking$value[3]` of 49</div>
<div style="font-size:0.8rem; color:#666;">tracked how many people were eligible</div>
</div>
<div style="flex:1 1 150px; min-width:150px; text-align:center; padding:0.95rem 0.5rem; background:#FDFBF7; border-radius:8px;">
<div style="font-size:1.8rem; font-weight:700; color:#B23A28;">`r notice$value[1]` of 50</div>
<div style="font-size:0.8rem; color:#666;">required that incarcerated people be told the policy exists</div>
</div>
</div>
<p class="chart-card-note">Tracking counts are from Holland et al. (2021),
<em>Mortality</em> 26(1), p. 55, across the forty-nine state systems that
responded to a public records request. The notification figure is from
Holland et al. (2020), <em>OMEGA</em> 81(4), p. 616: only Alabama and New
Mexico required that people in their custody be informed that the policy
existed.</p>
<p class="chart-card-note">Those two findings travel together: a policy
nobody is told about, tracked by nobody, will look underused whether or
not anyone would have used it.</p>
<p class="chart-card-note"><strong>Why this chart:</strong> tiles, because
four numbers with different denominators cannot share an axis, and forcing
them onto one would invent a comparison the data does not support. The
last tile is the one that changes how the first three read, so it goes
last.</p>
:::

## The Federal Docket {.section-heading}

::: {.chart-card}
<p class="chart-card-lede">Federal law runs on a different track: since
2018 people can file their own motions, and a judge rather than a prison
official decides. Here are four and a half years of that docket, month by
month.</p>
```{r chart-federal-monthly}
#| fig-height: 3.6
#| out-width: "100%"
p_fed <- ggplot(fed_month, aes(x = date)) +
  geom_col(aes(y = total, text = paste0(
    "<b>", label, "</b><br>",
    comma(total), " motions decided<br>",
    granted, " granted (", granted_pct, "%)"
  )), fill = "#e2dbe2", width = 26) +
  geom_col(aes(y = granted, text = paste0(
    "<b>", label, "</b><br>",
    granted, " granted of ", comma(total), " (", granted_pct, "%)"
  )), fill = plum_dark, width = 26) +
  scale_x_date(date_breaks = "1 year", date_labels = "%b %Y") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "The pandemic surge receded; the grant rate did not",
    subtitle = "Federal compassionate release motions decided each month, granted portion in dark",
    x = NULL, y = "Motions decided"
  ) +
  theme_cr()

ggplotly_titled(p_fed, tooltip = "text") |> clear_bg()
```
<p class="chart-card-note">Across the full period, courts decided 15,265
motions and granted 2,088 of them, or 13.7 percent. Monthly volume fell
from 586 in October 2021 to 140 in March 2026, while the share granted
rose from 9.2 percent to 16.4 percent. Source: U.S. Sentencing Commission,
Compassionate Release Data Reports: fiscal years 2022 through 2025 from
the Fiscal Year 2025 report, fiscal year 2026 from the 2nd Quarter FY2026
report, table 1 in each. Fiscal year 2026 figures are preliminary and
cumulative through the second quarter.</p>
<p class="chart-card-note"><strong>Why this chart:</strong> one bar nested
inside another, rather than two lines on two axes. Grants are a part of
motions decided, not a separate series, and drawing a part inside its whole
keeps the reader from reading a rising rate as a rising count. They moved
in opposite directions here, which is exactly the confusion a second axis
would have manufactured.</p>
:::

## Why Courts Say Yes, and Why They Say No {.section-heading}

::: {.chart-card}
<p class="chart-card-lede">Courts record their reasons, and the reasons are
the most direct evidence there is of what this law has come to mean in
practice. Grouped into themes, what does the federal docket say?</p>
```{r chart-reasons}
#| fig-height: 4.3
#| out-width: "100%"
rt <- reason_theme |>
  mutate(
    outcome_f = factor(outcome, levels = c("Granted", "Denied")),
    theme_f   = reorder(theme, share)
  )

p_rt <- ggplot(rt, aes(x = share, y = theme_f, fill = outcome_f)) +
  geom_col(aes(text = paste0(
    "<b>", theme, "</b><br>",
    outcome, ": ", comma(n), " reasons cited<br>",
    percent(share, accuracy = 0.1), " of reasons for that outcome"
  )), width = 0.68) +
  facet_wrap(~ outcome_f, ncol = 2) +
  scale_fill_manual(values = c(Granted = plum_series, Denied = coral_dark),
                    guide = "none") +
  scale_x_continuous(breaks = seq(0, 0.4, 0.2), limits = c(0, 0.47),
                     labels = function(x) 100 * x) +
  labs(
    title = "Sentence length wins release more often than illness does",
    subtitle = "Themes among reasons cited by federal courts, fiscal year 2025",
    x = "Percent of reasons cited for that outcome", y = NULL
  ) +
  theme_cr() +
  # Widen the gutter so the two facets' percent ticks stay apart at phone
  # widths.
  theme(panel.spacing.x = grid::unit(1.4, "lines"))

# No legend here on purpose: the facet strips already say Granted and Denied,
# so a legend would encode the same thing twice.
ggplotly_titled(p_rt, tooltip = "text") |> clear_bg()
```
<p class="chart-card-note">Among granted motions, reasons about sentence
length and changes in law accounted for 35.0 percent of citations and
medical reasons for 23.0 percent. Among denials, the single most cited
reason by far was the general sentencing factors at section 3553(a), a
discretionary weighing rather than a finding about health. Courts may cite
several reasons per case: 631 reasons across 394 granted cases, and 5,067
across 2,401 denied. Source: U.S. Sentencing Commission, Compassionate
Release Data Report, Fiscal Year 2025, tables 10 and 11.</p>
<p class="chart-card-note">One procedural line is worth reading on its
own. Failure to exhaust administrative remedies was cited 447 times, which
is a paperwork step rather than a judgment about whether release was
warranted.</p>
<p class="chart-card-note"><strong>Why this chart:</strong> small multiples
sharing one x scale, so grants and denials are read side by side without
either being subtracted from the other. They are not complements: a case
can be denied for four reasons and granted for two, so a single diverging
bar would have implied an arithmetic that does not exist.</p>
:::

## From Statute to Data {.section-heading}

:::: {.chart-card}
<p class="chart-card-lede">The chart above depends on a step that is easy
to skip past. The Sentencing Commission publishes reason codes; the themes
are mine, and the code labels are lightly shortened from the Commission's
wording. Here is the whole assignment, so the grouping can be argued
with.</p>
::: {.table-scroll}
```{r table-codebook}
#| output: asis
cb <- fed_reason |>
  arrange(outcome, theme, desc(n)) |>
  transmute(
    Outcome = outcome,
    `Theme (my coding)` = theme,
    `Reason code (the Commission's)` = reason,
    Cited = comma(n)
  )

knitr::kable(cb, align = c("l", "l", "l", "r"))
```
:::
<p class="chart-card-note">Two judgment calls in there are worth naming.
"Young age at time of offense" sits under sentence and law change rather
than under anything medical, because it is an argument about whether the
original sentence still fits. "Rehabilitation" is kept apart from both
health and sentence length, because the Commission notes it never appears
alone in a granted case: it always arrives alongside another reason. Three
codes are new in the full-year report: mandatory nature of the guideline
at sentencing, guideline amendment not yet in effect, and conviction or
sentencing errors. All three sit under sentence and law change, because
each is an argument about the sentence rather than the person's
condition.</p>
<p class="chart-card-note"><strong>Why this chart:</strong> a table, which
is the right chart when the reader's question is "what exactly did you
do." A chart of a coding scheme shows its shape; the scheme itself has to
be legible line by line, or nobody can tell you that you put a code in the
wrong pile.</p>
::::

## One State, Thirty Years {.section-heading}

::: {.chart-card}
<p class="chart-card-lede">National counts flatten a law into a yes or a
no. Any single state's version arrived in pieces, and Wisconsin's is a
clear case: a statute, a rollback, a rule, a set of forms, a policy.</p>
```{r chart-wi-timeline}
#| fig-height: 5.4
#| out-width: "100%"
wi_p <- wi |>
  mutate(
    lane = rep(c(1, -1), length.out = n()),
    kind_f = factor(kind, levels = c("Statute", "Rule", "Forms", "Policy")),
    lab = paste0(period, "\n", wrap_lab(event, 20))
  )

p_wi <- ggplot(wi_p, aes(x = 0, y = sort_year)) +
  geom_vline(xintercept = 0, color = "#dcd5dc", linewidth = 1.1) +
  geom_segment(aes(xend = lane * 0.05, yend = sort_year),
               color = "#dcd5dc", linewidth = 0.5) +
  geom_point(aes(text = paste0(
    "<b>", period, ": ", event, "</b><br>",
    detail, "<br>Source: ", source, " (", source_detail, ")"
  ), color = kind_f), size = 4.2) +
  geom_text(data = filter(wi_p, lane > 0),
            aes(x = 0.09, label = lab), hjust = 0,
            size = 3, color = "#3C3C3C", lineheight = 1.05) +
  geom_text(data = filter(wi_p, lane < 0),
            aes(x = -0.09, label = lab), hjust = 1,
            size = 3, color = "#3C3C3C", lineheight = 1.05) +
  scale_color_manual(values = c(Statute = plum_dark, Rule = plum_series,
                                Forms = gray_ctx, Policy = coral_dark)) +
  scale_x_continuous(limits = c(-1, 1), breaks = NULL) +
  scale_y_reverse(limits = c(2027, 1994), breaks = NULL) +
  labs(
    title = "A release mechanism assembled over twenty-seven years",
    subtitle = "Wisconsin medical, geriatric, and extraordinary health condition release, 1997 to 2024",
    x = NULL, y = NULL, color = NULL
  ) +
  theme_cr() +
  theme(axis.text = element_blank(), panel.grid.major = element_blank())

ggplotly_titled(p_wi, tooltip = "text") |>
  clear_bg() |>
  legend_below(y = -0.06, b = 70) |>
  anchor_text()
```
<p class="chart-card-note">Sources are the instruments themselves: 1997
Wis. Act 283, 2001 Wis. Act 109 creating Wis. Stat. sec. 302.113(9g),
2009 Wis. Act 28, 2011 Wis. Act 38, Wis. Admin. Code DOC sec. 302.41,
Wisconsin Court System forms CR-254 through CR-256, DAI Policy 302.00.13
(reissued effective June 22, 2026), and 2023 Wis. Act 230 on victim
statements. Today all four layers operate together and the sentencing
court makes the final call.</p>
<p class="chart-card-note">Notice the 2011 entry. A release mechanism does
not only get built; it gets narrowed, and a chart that recorded only
expansions would have shown a straight line up.</p>
<p class="chart-card-note"><strong>Why this chart:</strong> a timeline set
vertically, events alternating sides of the spine, because these labels
are phrases and phrases need horizontal room; along a horizontal axis each
label gets only the width of its own gap, and eight of them collide.
Color marks the kind of instrument, which is the point most easily
missed: what looks like one law is a statute, a rule, a form, and an
internal policy, each changeable by a different body on a different
schedule.</p>
:::

## What This Cannot Tell You {.section-heading}

Everything above is the law as written and the decisions as recorded. Both
are real, and both stop short of the question people actually care about,
which is why a law that exists so widely operates so rarely.

The records cannot answer it because the deciding happens before anything
gets written down. A prison official weighs whether to forward a petition.
A physician decides how confidently to put a prognosis in writing. A board
or a judge weighs a small, quiet benefit against a rare, loud risk. Those
are judgments made under uncertainty, and the file records the outcome, not
the reasoning that produced it.

That is where I think the useful work is. A few questions I would want
answered before claiming to know why these laws sit unused:

- Does what the policy is called change who applies and who is approved?
  The federal statute says "extraordinary and compelling." Most states have
  moved to medical or geriatric parole, medical release, or conditional
  medical release. Only New Jersey and Utah still use compassionate release
  as the exclusive title, and the words carry different promises: mercy in
  one, clinical necessity in the other.
- Does telling people the policy exists change the number who apply? Two
  states required it as of 2016, which is close enough to a natural
  experiment to be worth running.
- Do clearer medical criteria move decisions, or only move where the
  argument happens?
- When a deadline is written into the statute, does the decision arrive
  sooner, and does the effect survive the first few years?

None of those are answered here, and none of them can be answered from
statutes alone. They need the decisions, the timing, and the people who
made them.

Two boundaries are deliberate. The page does not mix the federal and state
figures into one national rate, because they count different things over
different periods under different law. And it does not report a denial
rate, because the published sources cannot support one: people leave these
processes by dying and by withdrawing, not only by being refused.

The state layer is the weak link, and it is weak in a specific way: it is
the best fifty-state coding anyone has published, and it describes a world
that has kept moving since 2016. Re-coding it from current statutes is the
next piece of work.

<p style="margin-top:2.2rem; font-size:0.95rem;">
Related on this site:
<a href="../howold/how_old_is_old.html">How Old Is Old?</a>, a short
exercise on putting numbers to a fuzzy category, and
<a href="../countedwrong/index.html">Maturity Gap</a>, which asks the same
kind of question at the other end of a life.
</p>
