---
title: "Data Dictionary & Codebook"
subtitle: "Hip-Hop Periodic Table: what the variables mean, how scores were assigned, and who is in the data"
format:
  html:
    toc: true
    toc-depth: 2
---

```{r setup}
#| include: false
library(readxl)
library(dplyr)
library(reactable)

dict_raw <- read_excel("data/hiphop_artists.xlsx", sheet = "Data Dictionary")

# The sheet has the variable table first, then a "SAMPLING FRAME & SCOPE RULES"
# block, then a "STYLE TAXONOMY" block; split on the markers so each renders
# in its own section.
marker  <- which(dict_raw[[1]] == "SAMPLING FRAME & SCOPE RULES")
smarker <- which(dict_raw[[1]] == "STYLE TAXONOMY")
vars  <- dict_raw[seq_len(marker - 1), ] |> filter(!is.na(Variable))
frame <- dict_raw[(marker + 1):(smarker - 1), 1:2] |>
  setNames(c("Rule", "Definition")) |> filter(!is.na(Rule))
style_rule <- dict_raw[[2]][smarker]
styles <- dict_raw[(smarker + 1):nrow(dict_raw), 1:2] |>
  setNames(c("Style", "Definition")) |> filter(!is.na(Style))
```

This page documents every variable in the [Hip-Hop Periodic
Table](hiphop_periodic_table.html) dataset, the constructs behind the five
lyricism scores, how scores were assigned, and the sampling frame the dataset
draws from. The same content ships as the **Data Dictionary** sheet inside the
Excel workbook, which is the single source of truth.

## The five constructs

Each act is scored 1–10 on five dimensions of lyricism. Scores are
**drawn from published criticism and hip-hop scholarship**; no automated
text analysis was run. Each dimension has a named anchor at 10 so the top
of the scale means the same thing across scorers.

| Construct | What it measures | Anchor at 10 |
|---|---|---|
| **Rhyme Density** | Frequency and complexity of rhyme: internal rhyme, multisyllabic chains, rhymes per bar, scheme inventiveness | Rakim: pioneered internal and multisyllabic rhyme as standard practice |
| **Vocab Breadth** | Lexical diversity: range of unique words and registers across the catalog | Aesop Rock: per published studies counting unique words in rap |
| **Storytelling** | Narrative construction: character, plot, perspective shifts, arc within songs and across albums | Scarface |
| **Metaphor/Imagery** | Density and originality of figurative language: extended metaphor, concrete imagery, double meanings | Lil Wayne: simile machine; metaphor-stacking as his defining critical reputation |
| **Conceptual Depth** | Thematic ambition and coherence: album-level concepts, social and philosophical substance | Kendrick Lamar: 10 is reserved for carrying a concept across a full album at the highest level |

The **Composite Score** is the plain average of the five dimensions, rounded
to one decimal. Counting all five equally is itself a choice, and one of the
module's core discussion prompts.

## Scoring method

- **Basis:** published music criticism, academic hip-hop scholarship
  (Bradley, Chang, Kajikawa, Rabaka), and documented critical consensus.
  Scores are judgments, not calculations.
- **Scale:** integers 1–10 per dimension; anchored at the top of each scale
  (table above).
- **Confidence flag:** every act carries an H/M/L rating of how much to trust
  its scores, based on the depth of its critical record (rubric below).

## Confidence rubric

| Flag | Meaning | Criteria |
|---|---|---|
| **H (High)** | Extensive critical record | Sustained critical/academic attention; multiple long-form analyses of the act's lyrical craft; consistent reception across sources; a body of work large enough that scores are stable across albums |
| **M (Medium)** | Moderate documentation | Meets 1–2 of the High criteria: historically important but under-analyzed, newer with an incomplete body of work, mixed critical reception, or regional/underground importance without mainstream scholarly attention. Treat as a working estimate |
| **L (Low)** | Thin or disputed record | Little formal analysis, a very recent debut, single-source scores, or reception clouded by controversy unrelated to the music. Treat as first estimates that need checking |

Good places to double-check M and L scores: RateYourMusic critical consensus,
Pitchfork / The Wire discography reviews, Google Scholar (artist +
"lyricism"), and genre publications (The Source, XXL retrospectives, Passion
of the Weiss).

## Sampling frame & scope rules

```{r frame-table}
#| echo: false
reactable(
  frame,
  columns = list(
    Rule = colDef(minWidth = 130, style = list(fontWeight = "600")),
    Definition = colDef(minWidth = 380)
  ),
  striped = TRUE, bordered = TRUE, sortable = FALSE, pagination = FALSE,
  theme = reactableTheme(
    headerStyle = list(background = "#2a1f2b", color = "#ffffff", fontWeight = "600")
  )
)
```

Two scope rules are worth calling out:

- **Scene = scene, not birthplace.** Acts are assigned to the US scene they
  are professionally rooted in: label home and scene ties during their
  defining run (Da Brat is *South* via So So Def, though born in Chicago).
  **Region** rolls scenes up to the four coast-level categories the culture
  itself uses: East Coast (NYC + Philly), West Coast (LA + Bay Area), South,
  and Midwest.
- **Era = the signature-year bracket.** Era is derived, not judged: it is the
  bracket holding the Signature Work year (Debut Year only when no signature
  exists, i.e. Kool Herc). Brackets: Old School through 1985, Golden Age
  1986–1994, Late 90s 1995–1999, then calendar decades. Debut years still
  overlap era boundaries by design; the judgment lives in the Signature Work
  pick, which has its own documented procedure (see Variable definitions).

## The eight styles

Style is one value per act: the dominant mode of the act's defining run,
judged from critical descriptions of the catalog. The axis separating the two
left-field categories is where the strangeness lives: in the words, it is
Abstract; in the sound or form, it is Experimental. Each definition names an
anchor act, the same way the five constructs anchor their scales.

```{r styles-table}
#| echo: false
reactable(
  styles,
  columns = list(
    Style = colDef(minWidth = 130, style = list(fontWeight = "600")),
    Definition = colDef(minWidth = 420)
  ),
  striped = TRUE, bordered = TRUE, sortable = FALSE, pagination = FALSE,
  theme = reactableTheme(
    headerStyle = list(background = "#2a1f2b", color = "#ffffff", fontWeight = "600")
  )
)
```

## Variable definitions

```{r vars-table}
#| echo: false
reactable(
  vars,
  columns = list(
    Variable = colDef(minWidth = 110, style = list(fontWeight = "600")),
    Type = colDef(minWidth = 80),
    `Allowed values / range` = colDef(minWidth = 160),
    Definition = colDef(minWidth = 340),
    `Source / method` = colDef(minWidth = 200)
  ),
  striped = TRUE, bordered = TRUE, sortable = FALSE, pagination = FALSE,
  theme = reactableTheme(
    headerStyle = list(background = "#2a1f2b", color = "#ffffff", fontWeight = "600")
  )
)
```

## Contested calls

The Signature Work column drives Era, so it carries real weight, and for a
handful of acts the critical record genuinely splits between two candidate
releases. Those calls are not hidden: the 2026-07 audit reviewed every pick
against the documented procedure, changed ten where the consensus clearly
pointed elsewhere, and resolved the genuine splits by maintainer ruling. The
rulings are logged here so the judgment is inspectable.

| Act | The split | Ruling | Why |
|---|---|---|---|
| Common | *Resurrection* (1994) vs *Like Water for Chocolate* (2000) vs *Be* (2005) | *Resurrection* | The album 90s critics call his best work, and the earliest of a three-way split |
| 2Pac | *Me Against the World* (1995) vs *All Eyez on Me* (1996) | *All Eyez on Me* | The cultural monument outweighs the critics' cohesion pick |
| Busta Rhymes | *The Coming* (1996) vs *When Disaster Strikes* (1997) | *When Disaster Strikes* | The peak-era statement over the classic debut |
| Royce da 5'9" | *Death Is Certain* (2004) vs *Book of Ryan* (2018) | *Book of Ryan* | The critical high point of the catalog |
| 8Ball & MJG | *Comin' Out Hard* (1993) vs *On Top of the World* (1995) | *On Top of the World* | Their highest-rated album over the influence landmark |
| Roc Marciano | *Marcberg* (2010) vs *Reloaded* (2012) | *Reloaded* | The refined peak edges the blueprint on critic scores |
| Little Brother | *The Listening* (2003) vs *The Minstrel Show* (2005) | *The Minstrel Show* | The fan-canon favorite |
| Xzibit | *At the Speed of Life* (1996) vs *40 Dayz & 40 Nightz* (1998) | *40 Dayz & 40 Nightz* | The album most often called his best |

If you would rule differently on any of these, that disagreement is the
point: the pick is documented, so the argument has something to grab.

## Known biases

The dataset over-represents critically acclaimed acts (vs. commercially
dominant ones), male acts, Golden Age NYC, and English-language rap.
International and non-English rap is **excluded from the frame entirely**,
a documented scope decision. The [dashboard](hiphop_dashboard.html) section
*"Who Is In the Data?"* measures these gaps and turns them into teaching
prompts on sampling bias.
