---
title: "Data Dictionary & Codebook"
subtitle: "Hip-Hop Periodic Table: what the variables mean, how scores were assigned, and who is in the data"
format:
  html:
    toc: true
    toc-depth: 2
---

```{r setup}
#| include: false
library(readxl)
library(dplyr)
library(reactable)

dict_raw <- read_excel("data/hiphop_artists.xlsx", sheet = "Data Dictionary")

# The sheet has the variable table first, then a "SAMPLING FRAME & SCOPE RULES"
# block; split on that marker so each renders in its own section.
marker <- which(dict_raw[[1]] == "SAMPLING FRAME & SCOPE RULES")
vars  <- dict_raw[seq_len(marker - 1), ] |> filter(!is.na(Variable))
frame <- dict_raw[(marker + 1):nrow(dict_raw), 1:2] |>
  setNames(c("Rule", "Definition")) |> filter(!is.na(Rule))
```

This page documents every variable in the [Hip-Hop Periodic
Table](hiphop_periodic_table.html) dataset, the constructs behind the five
lyricism scores, how scores were assigned, and the sampling frame the dataset
draws from. The same content ships as the **Data Dictionary** sheet inside the
Excel workbook, which is the single source of truth.

## The five constructs

Each act is scored 1–10 on five dimensions of lyricism. Scores are
**drawn from published criticism and hip-hop scholarship**; no automated
text analysis was run. Each dimension has a named anchor at 10 so the top
of the scale means the same thing across scorers.

| Construct | What it measures | Anchor at 10 |
|---|---|---|
| **Rhyme Density** | Frequency and complexity of rhyme: internal rhyme, multisyllabic chains, rhymes per bar, scheme inventiveness | Rakim: pioneered internal and multisyllabic rhyme as standard practice |
| **Vocab Breadth** | Lexical diversity: range of unique words and registers across the catalog | Aesop Rock: per published studies counting unique words in rap |
| **Storytelling** | Narrative construction: character, plot, perspective shifts, arc within songs and across albums | Scarface |
| **Metaphor/Imagery** | Density and originality of figurative language: extended metaphor, concrete imagery, double meanings | (none named) |
| **Conceptual Depth** | Thematic ambition and coherence: album-level concepts, social and philosophical substance | Kendrick Lamar: 10 is reserved for carrying a concept across a full album at the highest level |

The **Composite Score** is the plain average of the five dimensions, rounded
to one decimal. Counting all five equally is itself a choice, and one of the
module's core discussion prompts.

## Scoring method

- **Basis:** published music criticism, academic hip-hop scholarship
  (Bradley, Chang, Kajikawa, Rabaka), and documented critical consensus.
  Scores are judgments, not calculations.
- **Scale:** integers 1–10 per dimension; anchored at the top of each scale
  (table above).
- **Confidence flag:** every act carries an H/M/L rating of how much to trust
  its scores, based on the depth of its critical record (rubric below).

## Confidence rubric

| Flag | Meaning | Criteria |
|---|---|---|
| **H (High)** | Extensive critical record | Sustained critical/academic attention; multiple long-form analyses of the act's lyrical craft; consistent reception across sources; a body of work large enough that scores are stable across albums |
| **M (Medium)** | Moderate documentation | Meets 1–2 of the High criteria: historically important but under-analyzed, newer with an incomplete body of work, mixed critical reception, or regional/underground importance without mainstream scholarly attention. Treat as a working estimate |
| **L (Low)** | Thin or disputed record | Little formal analysis, a very recent debut, single-source scores, or reception clouded by controversy unrelated to the music. Treat as first estimates that need checking |

Good places to double-check M and L scores: RateYourMusic critical consensus,
Pitchfork / The Wire discography reviews, Google Scholar (artist +
"lyricism"), and genre publications (The Source, XXL retrospectives, Passion
of the Weiss).

## Sampling frame & scope rules

```{r frame-table}
#| echo: false
reactable(
  frame,
  columns = list(
    Rule = colDef(minWidth = 130, style = list(fontWeight = "600")),
    Definition = colDef(minWidth = 380)
  ),
  striped = TRUE, bordered = TRUE, sortable = FALSE, pagination = FALSE,
  theme = reactableTheme(
    headerStyle = list(background = "#1F3864", color = "#ffffff", fontWeight = "600")
  )
)
```

Two scope rules are worth calling out:

- **Region = scene, not birthplace.** Acts are assigned to the US scene they
  are professionally rooted in: label home and scene ties during their
  defining run (Da Brat is *South* via So So Def, though born in Chicago).
- **Era = defining run, not debut year.** Debut years overlap era boundaries
  by design; the era reflects when the act's defining work landed.

## Variable definitions

```{r vars-table}
#| echo: false
reactable(
  vars,
  columns = list(
    Variable = colDef(minWidth = 110, style = list(fontWeight = "600")),
    Type = colDef(minWidth = 80),
    `Allowed values / range` = colDef(minWidth = 160),
    Definition = colDef(minWidth = 340),
    `Source / method` = colDef(minWidth = 200)
  ),
  striped = TRUE, bordered = TRUE, sortable = FALSE, pagination = FALSE,
  theme = reactableTheme(
    headerStyle = list(background = "#1F3864", color = "#ffffff", fontWeight = "600")
  )
)
```

## Known biases

The dataset over-represents critically acclaimed acts (vs. commercially
dominant ones), male acts, Golden Age NYC, and English-language rap.
International and non-English rap is **excluded from the frame entirely**,
a documented scope decision. The [dashboard](hiphop_dashboard.html) section
*"Who Is In the Data?"* measures these gaps and turns them into teaching
prompts on sampling bias.
