---
title: "Hip-Hop Periodic Table"
subtitle: "A Data Storytelling Teaching Module"
author: "Data Storytelling in Professional Practice"
format:
  dashboard:
    theme: 
      - cosmo
      - ../custom.scss
    logo: ""
    nav-buttons:
      - icon: github
        href: "https://github.com/lwylie01/my-site"
    scrolling: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(scales)
library(forcats)
library(stringr)
library(reactable)
library(htmltools)
library(readxl)

# ── Load data from Excel ───────────────────────────────────────────────────────
# File lives at: my-site/hiphop/data/hiphop_artists.xlsx
# To add new artists: open the Excel file, append rows to the "Artist Database"
# sheet, then push to main — the publish.yml GitHub Action rebuilds the site.

artists_raw <- read_excel(
  "data/hiphop_artists.xlsx",
  sheet = "Artist Database"
) |>
  # The dataset mixes individual artists ("Element") with groups ("Compound",
  # e.g. Wu-Tang Clan). This dashboard analyses individuals only; groups are
  # shown separately in the periodic table.
  filter(Type == "Element") |>
  rename(
    sym            = Symbol,
    name           = `Artist Name`,
    region         = Region,
    era            = Era,
    style          = Style,
    debut          = `Debut Year`,
    rhyme          = `Rhyme Density`,
    vocab          = `Vocab Breadth`,
    story          = Storytelling,
    metaphor       = `Metaphor/Imagery`,
    concept        = `Conceptual Depth`,
    confidence     = Confidence,
    confidence_note = `Confidence Note`,
    gender         = Gender,
    sig_work       = `Signature Work`,
    active_to      = `Active Through`,
    birth          = `Birth Year`
  ) |>
  mutate(across(c(rhyme, vocab, story, metaphor, concept), as.integer))

# ── Derived variables ─────────────────────────────────────────────────────────
current_year <- as.integer(format(Sys.Date(), "%Y"))

artists <- artists_raw |>
  mutate(
    composite = round((rhyme + vocab + story + metaphor + concept) / 5, 1),
    career_end    = coalesce(as.integer(active_to), current_year),
    career_len    = career_end - debut,
    sig_year      = as.integer(str_match(sig_work, "\\((\\d{4})\\)")[, 2]),
    years_to_peak = sig_year - debut,
    age_debut     = debut - birth,
    age_sig       = sig_year - birth,
    era = factor(era, levels = c("Old School","Golden Age","Late 90s","2000s","2010s","2020s")),
    confidence = factor(confidence, levels = c("H","M","L")),
    region = factor(region, levels = c("East Coast","West Coast","South","Midwest")),
    style = factor(style),
    confidence_note = if_else(is.na(confidence_note), "", confidence_note),
    strongest_dim = case_when(
      rhyme == pmax(rhyme, vocab, story, metaphor, concept) ~ "Rhyme Density",
      vocab == pmax(rhyme, vocab, story, metaphor, concept) ~ "Vocab Breadth",
      story == pmax(rhyme, vocab, story, metaphor, concept) ~ "Storytelling",
      metaphor == pmax(rhyme, vocab, story, metaphor, concept) ~ "Metaphor/Imagery",
      TRUE ~ "Conceptual Depth"
    )
  )

# ── Colour palettes ──────────────────────────────────────────────────────────
# Brand-family categorical hues, borrowed from the periodic table's tile and
# score-bar palettes so the two pages read as one product.
region_pal <- c(
  "East Coast" = "#6f9ec2",
  "West Coast" = "#cf8079",
  "South"      = "#86a866",
  "Midwest"    = "#a085bd"
)

era_pal <- c(
  "Old School" = "#5a9abf",
  "Golden Age" = "#d4b15a",
  "Late 90s"   = "#86a866",
  "2000s"      = "#dd9a5e",
  "2010s"      = "#a085bd",
  "2020s"      = "#cf8079"
)

# Same five hues as the periodic table's modal score bars (SCORE_HUES), so
# each dimension keeps one color across the whole site.
dim_pal <- c(
  "Rhyme Density"    = "#E98973",
  "Vocab Breadth"    = "#8C6A8D",
  "Storytelling"     = "#7aaa8a",
  "Metaphor/Imagery" = "#5a9abf",
  "Conceptual Depth" = "#bf9a40"
)

conf_pal <- c("H" = "#41ab5d", "M" = "#f6c141", "L" = "#e08214")

# ── Shared ggplot2 theme ─────────────────────────────────────────────────────
theme_hiphop <- function() {
  theme_minimal(base_family = "sans", base_size = 12) +
    theme(
      plot.title    = element_text(size = 14, face = "bold", color = "#3C3C3C"),
      plot.subtitle = element_text(size = 11, color = "#555555"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#eeeeee"),
      axis.text     = element_text(color = "#444444"),
      legend.position = "bottom",
      legend.title  = element_text(size = 10, face = "bold"),
      strip.text    = element_text(face = "bold", color = "#333333")
    )
}

# ── ggplotly() drops ggplot subtitles ────────────────────────────────────────
# Fold each subtitle into the plotly title as a smaller <br><sup> second line
# so it survives the conversion. The <b> wrapper matches the bold title text
# ggplotly emits; the margin bump keeps both lines clear of the plot area.
# Pipe every ggplot chart through this instead of bare ggplotly().
ggplotly_titled <- function(p, ...) {
  fig <- ggplotly(p, ...)
  sub <- p$labels$subtitle
  if (is.null(sub) || !nzchar(sub)) return(fig)
  fig |>
    layout(
      title = list(text = paste0(
        "<b>", p$labels$title, "</b><br><sup>", sub, "</sup>"
      )),
      margin = list(t = 70)
    )
}
```

# Overview {.sidebar}

```{r sidebar-summary}
n_artists  <- nrow(artists)
n_high     <- sum(artists$confidence == "H")
n_medium   <- sum(artists$confidence == "M")
n_low      <- sum(artists$confidence == "L")
avg_comp   <- round(mean(artists$composite), 2)
top_artist <- artists |> slice_max(composite, n = 1) |> pull(name)

tagList(
  tags$div(style = "padding: 1rem 0;",
    tags$h4("Dataset at a Glance", style = "color:#3C3C3C; font-weight:700;"),
    tags$hr(style = "border-color:#dddddd;"),
    tags$p(tags$strong(n_artists), " artists across 6 eras"),
    tags$p(tags$strong(length(levels(artists$region))), " geographic regions"),
    tags$p(tags$strong(length(levels(artists$style))), " lyrical styles"),
    tags$hr(style = "border-color:#dddddd;"),
    tags$h5("Confidence breakdown", style = "color:#555;"),
    tags$p(
      tags$span(style = "color:#41ab5d; font-weight:700;", paste0("H: ", n_high, " artists")),
      tags$br(),
      tags$span(style = "color:#c8a000; font-weight:700;", paste0("M: ", n_medium, " artists (some gaps)")),
      tags$br(),
      tags$span(style = "color:#e08214; font-weight:700;", paste0("L: ", n_low, " artists (estimates only)"))
    ),
    tags$hr(style = "border-color:#dddddd;"),
    tags$h5("Mean composite score", style = "color:#555;"),
    tags$p(tags$strong(avg_comp, " / 10")),
    tags$p(tags$em(paste0("Highest: ", top_artist))),
    tags$hr(style = "border-color:#dddddd;"),
    tags$h5("Teaching concepts", style = "color:#555; margin-top:1rem;"),
    tags$ul(style = "padding-left:1.2rem; font-size:0.85rem;",
      tags$li("Measurement validity"),
      tags$li("Sampling & bias"),
      tags$li("Multidimensional scoring"),
      tags$li("Composite vs. dimensions"),
      tags$li("Uncertainty communication"),
      tags$li("Visualization design")
    )
  )
)
```

# 1 · What Are We Measuring?

## Row {height="15%"}

::: {.callout-tip title="Teaching Prompt: Construct Validity"}
**The central measurement problem:** We started with one variable called "Lyricism" and ended up with five. Was that the right decision? What are we gaining and losing? This is the construct validity question that underlies every performance measure, survey instrument, and rating scale in your organization.
:::

## Row {height="85%"}

### Radar: One artist, five dimensions {width="40%"}

```{r radar-single}
radar_artists <- c("Kendrick Lamar", "Scarface", "MF DOOM", "Rakim")

radar_data <- artists |>
  filter(name %in% radar_artists) |>
  select(name, rhyme, vocab, story, metaphor, concept) |>
  pivot_longer(-name, names_to = "dimension", values_to = "score") |>
  mutate(dimension = recode(dimension,
    rhyme    = "Rhyme Density",
    vocab    = "Vocab Breadth",
    story    = "Storytelling",
    metaphor = "Metaphor/Imagery",
    concept  = "Conceptual Depth"
  ))

fig_radar <- plot_ly(type = "scatterpolar", fill = "toself", mode = "lines+markers")

colors_radar <- c("#8C6A8D","#E98973","#5a9abf","#bf9a40")
for (i in seq_along(radar_artists)) {
  d <- radar_data |> filter(name == radar_artists[i])
  fig_radar <- fig_radar |> add_trace(
    r = c(d$score, d$score[1]),
    theta = c(d$dimension, d$dimension[1]),
    name = radar_artists[i],
    line = list(color = colors_radar[i], width = 2),
    fillcolor = paste0(colors_radar[i], "22")
  )
}

fig_radar |>
  layout(
    polar = list(radialaxis = list(visible = TRUE, range = c(0, 10), tickfont = list(size = 10))),
    legend = list(orientation = "h", x = 0.1, y = -0.1),
    margin = list(t = 10, b = 10),
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)"
  )
```

### Scatter: Composite score vs. Storytelling {width="60%"}

```{r scatter-composite-v-story}
p_scatter <- artists |>
  filter(confidence != "L") |>
  ggplot(aes(x = story, y = composite,
             color = era, size = composite,
             text = paste0(
               "<b>", name, "</b><br>",
               "Era: ", era, "<br>",
               "Composite: ", composite, "<br>",
               "Storytelling: ", story, "<br>",
               "Rhyme Density: ", rhyme
             ))) +
  geom_jitter(alpha = 0.75, width = 0.15, height = 0.05) +
  scale_color_manual(values = era_pal) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(2, 10.5), breaks = seq(2, 10, 2)) +
  labs(
    title = "Composite score vs. Storytelling dimension",
    subtitle = "Artists with the same composite can have very different dimension profiles",
    x = "Storytelling score (1–10)",
    y = "Composite score",
    color = "Era"
  ) +
  theme_hiphop() +
  theme(legend.position = "right")

ggplotly_titled(p_scatter, tooltip = "text") |>
  layout(
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)"
  )
```

# 2 · Who Is In the Data?

## Row {height="15%"}

::: {.callout-warning title="Teaching Prompt: Sampling Decisions and Bias"}
**Every dataset has a frame, and every frame has a gap.** This dataset over-represents critically acclaimed artists, the East Coast (NYC especially), the Golden Age, and men. Those are documented choices. What datasets do you use at work where nobody wrote the frame down? Who is missing?
:::

## Row {height="45%"}

### Artists by region {width="50%"}

```{r bar-region}
p_region <- artists |>
  count(region, era) |>
  ggplot(aes(x = fct_rev(region), y = n, fill = era,
             text = paste0(era, ": ", n, " artists"))) +
  geom_col(position = "stack") +
  scale_fill_manual(values = era_pal) +
  coord_flip() +
  labs(
    title = "Artists by region and era",
    subtitle = "The East Coast and South dominate; international scenes sit outside the frame",
    x = NULL, y = "Number of artists", fill = "Era"
  ) +
  theme_hiphop()

ggplotly_titled(p_region, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### Gender representation gap {width="50%"}

```{r bar-gender}
# Gender comes from the Gender column in the Excel workbook (single source of
# truth). "Mixed" only occurs for Compounds, which this dashboard filters out.
gender_data <- artists |>
  count(era, gender_group = gender)

n_female <- sum(artists$gender == "Female")

p_gender <- gender_data |>
  ggplot(aes(x = era, y = n, fill = gender_group,
             text = paste0(gender_group, ": ", n, " artists"))) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c("Female" = "#c180a2", "Male" = "#b3abb3")) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Gender representation by era",
    subtitle = paste0(
      n_female, " female artists of ", nrow(artists), " total (",
      round(n_female/nrow(artists)*100), "%)"
    ),
    x = NULL, y = "Share of era", fill = NULL
  ) +
  theme_hiphop()

ggplotly_titled(p_gender, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

## Row {height="40%"}

### Confidence distribution {width="35%"}

```{r confidence-donut}
conf_counts <- artists |>
  count(confidence) |>
  mutate(
    label = case_when(
      confidence == "H" ~ paste0("High\n", n, " artists"),
      confidence == "M" ~ paste0("Medium\n", n, " artists"),
      confidence == "L" ~ paste0("Low\n", n, " artists")
    )
  )

plot_ly(conf_counts,
  labels = ~label, values = ~n,
  type = "pie", hole = 0.55,
  marker = list(colors = c("#41ab5d","#f6c141","#e08214"),
                line = list(color = "#ffffff", width = 2)),
  textinfo = "label+percent",
  hoverinfo = "label+value"
) |>
  layout(
    title = list(text = "Scoring confidence distribution", font = list(size = 13)),
    showlegend = FALSE,
    paper_bgcolor = "rgba(0,0,0,0)"
  )
```

### What bias looks like in practice {width="65%"}

```{r bias-table}
bias_df <- tribble(
  ~"Bias type",        ~"Direction",                              ~"Practical implication",
  "Prestige",         "Over-represents critically acclaimed artists", "Commercial artists scored lower; vibe-first artists may be underrated",
  "Gender",           "~18% female vs. ~50% of rap audience",    "Female artists face different critical frameworks; scores may not be comparable",
  "Geography",        "East Coast + Golden Age disproportionate", "Southern and international artists underrepresented in critical canon used for scoring",
  "Language",         "English-language, US-based frame only",    "International and non-English rap is excluded from the sampling frame entirely, a documented scope decision rather than an oversight",
  "Survivorship",     "Only artists who entered critical canon",  "Regionally significant artists without national coverage are absent",
  "Recency (inverted)","Post-2020 artists scored conservatively", "Emerging talent may be systematically underscored; revisit annually"
)

reactable(
  bias_df,
  columns = list(
    "Bias type"             = colDef(minWidth = 100, style = list(fontWeight = "600")),
    "Direction"             = colDef(minWidth = 160),
    "Practical implication" = colDef(minWidth = 200)
  ),
  striped = TRUE, highlight = TRUE, bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(background = "#2a1f2b", color = "#ffffff", fontWeight = "600")
  )
)
```

# 3 · Dimensions vs. Composite

## Row {height="15%"}

::: {.callout-note title="Teaching Prompt: The Averaging Problem"}
**What does averaging destroy?** Rakim (Rhyme Density 10, composite 8.8) and Scarface (Storytelling 10, composite 8.2) land within half a point of each other on the composite, yet their profiles look nothing alike. Does that similarity mean anything, or does the average erase the most important information? When does averaging help, and when does it mislead?
:::

## Row {height="45%"}

### Heatmap: All artists, all dimensions {width="65%"}

```{r heatmap-dims}
dim_long <- artists |>
  filter(confidence == "H") |>
  arrange(era, composite) |>
  select(name, era, rhyme, vocab, story, metaphor, concept) |>
  pivot_longer(rhyme:concept, names_to = "dimension", values_to = "score") |>
  mutate(
    dimension = recode(dimension,
      rhyme    = "Rhyme",
      vocab    = "Vocab",
      story    = "Story",
      metaphor = "Metaphor",
      concept  = "Concept"
    ),
    dimension = factor(dimension, levels = c("Rhyme","Vocab","Story","Metaphor","Concept")),
    name = factor(name, levels = unique(name))
  )

p_heat <- dim_long |>
  ggplot(aes(x = dimension, y = name, fill = score,
             text = paste0(name, " · ", dimension, ": ", score))) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low = "#f6f0f6", mid = "#a87ea9", high = "#4a2d4b",
    midpoint = 5.5, limits = c(1, 10),
    name = "Score"
  ) +
  facet_grid(era ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Dimension scores by artist (High-confidence only)",
    subtitle = "Each column is a dimension; each row an artist; color = score 1–10",
    x = NULL, y = NULL
  ) +
  theme_hiphop() +
  theme(
    axis.text.y = element_text(size = 7),
    strip.text.y = element_text(angle = 0, size = 8),
    panel.spacing = unit(0.3, "lines")
  )

ggplotly_titled(p_heat, tooltip = "text", height = 900) |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### Era averages by dimension {width="35%"}

```{r era-dim-lines}
era_dim_avg <- artists |>
  group_by(era) |>
  summarise(
    `Rhyme Density`    = mean(rhyme),
    `Vocab Breadth`    = mean(vocab),
    `Storytelling`     = mean(story),
    `Metaphor/Imagery` = mean(metaphor),
    `Conceptual Depth` = mean(concept),
    .groups = "drop"
  ) |>
  pivot_longer(-era, names_to = "dimension", values_to = "avg")

p_era_lines <- era_dim_avg |>
  ggplot(aes(x = era, y = avg, color = dimension, group = dimension,
             text = paste0(dimension, " (", era, "): ", round(avg, 1)))) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = dim_pal) +
  scale_y_continuous(limits = c(4, 9), breaks = seq(4, 9, 1)) +
  labs(
    title = "Dimension averages across eras",
    subtitle = "Which dimensions rose or fell over time?",
    x = NULL, y = "Mean score", color = NULL
  ) +
  theme_hiphop() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly_titled(p_era_lines, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

## Row {height="40%"}

### Strongest dimension by style {width="50%"}

```{r strongest-dim-style}
strongest <- artists |>
  count(style, strongest_dim) |>
  group_by(style) |>
  mutate(pct = n / sum(n))

p_strongest <- strongest |>
  ggplot(aes(x = style, y = pct, fill = strongest_dim,
             text = paste0(strongest_dim, ": ", round(pct*100), "%"))) +
  geom_col(position = "fill") +
  scale_fill_manual(values = dim_pal) +
  scale_y_continuous(labels = percent) +
  coord_flip() +
  labs(
    title = "Dominant dimension by style",
    subtitle = "What does each style specialize in?",
    x = NULL, y = "Share", fill = "Strongest dimension"
  ) +
  theme_hiphop() +
  theme(legend.position = "bottom")

ggplotly_titled(p_strongest, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### Top 20 composite scores (High confidence) {width="50%"}

```{r top20-bar}
top20 <- artists |>
  filter(confidence == "H") |>
  slice_max(composite, n = 20) |>
  arrange(composite) |>
  mutate(name = factor(name, levels = name))

p_top20 <- top20 |>
  ggplot(aes(x = name, y = composite, fill = era,
             text = paste0(name, "\n", era, " | ", style, "\nComposite: ", composite))) +
  geom_col() +
  scale_fill_manual(values = era_pal) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 10.5)) +
  labs(
    title = "Top 20 composite scores (High confidence)",
    subtitle = "Composite = unweighted average of 5 dimensions",
    x = NULL, y = "Composite score (1–10)", fill = "Era"
  ) +
  theme_hiphop()

ggplotly_titled(p_top20, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

# 4 · Uncertainty in Practice

## Row {height="15%"}

::: {.callout-important title="Teaching Prompt: Communicating Uncertainty"}
**Most dashboards show you the number. This one shows you how much to trust it.** The confidence flag almost never shows up in stakeholder reporting, but it should. Where in your organization's reporting does uncertainty get hidden? What decisions are being made on L-confidence data presented as H-confidence?
:::

## Row {height="45%"}

### Composite distribution by confidence tier {width="50%"}

```{r violin-confidence}
violin_data <- artists |>
  mutate(
    conf_label = factor(
      case_when(confidence == "H" ~ "High", confidence == "M" ~ "Medium", TRUE ~ "Low"),
      levels = c("High", "Medium", "Low")
    ),
    composite = as.numeric(composite)
  )

plot_ly(
  violin_data,
  x        = ~conf_label,
  y        = ~composite,
  split    = ~conf_label,
  type     = "violin",
  text     = ~paste0(name, ": ", composite),
  hoverinfo = "text",
  points   = "all",
  jitter   = 0.3,
  pointpos = 0,
  opacity  = 0.6,
  marker   = list(size = 5, opacity = 0.8),
  color    = ~conf_label,
  colors   = c("High" = "#41ab5d", "Medium" = "#f6c141", "Low" = "#e08214"),
  showlegend = FALSE
) |>
  layout(
    title  = list(text = "Score distribution by confidence tier",
                  font = list(size = 14, color = "#3C3C3C")),
    xaxis  = list(title = "Confidence tier"),
    yaxis  = list(title = "Composite score", range = c(1, 10.5)),
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)"
  )
```

### Score range: What we don't know {width="50%"}

```{r uncertainty-range}
uncertain <- artists |>
  filter(confidence %in% c("M","L")) |>
  arrange(desc(composite)) |>
  head(25) |>
  mutate(
    lo = pmax(1,  composite - if_else(confidence == "M", 0.8, 1.6)),
    hi = pmin(10, composite + if_else(confidence == "M", 0.8, 1.6)),
    name = fct_reorder(name, composite)
  )

p_range <- uncertain |>
  ggplot(aes(y = name, color = confidence,
             text = paste0(name, "\nScore: ", composite,
                           "\nPlausible range: ", round(lo,1), "–", round(hi,1)))) +
  geom_segment(aes(x = lo, xend = hi, yend = name), linewidth = 1.8, alpha = 0.4) +
  geom_point(aes(x = composite), size = 3) +
  scale_color_manual(values = conf_pal,
                     labels = c("M" = "Medium (±0.8)", "L" = "Low (±1.6)")) +
  scale_x_continuous(limits = c(1, 10), breaks = 1:10) +
  labs(
    title = "Plausible score ranges for less-documented artists",
    subtitle = "Point = scored value; bar = plausible range given evidence quality",
    x = "Composite score", y = NULL, color = "Confidence tier"
  ) +
  theme_hiphop()

ggplotly_titled(p_range, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

## Row {height="40%"}

### Confidence notes for M and L entries {width="100%"}

```{r conf-notes-table}
conf_notes <- artists |>
  filter(confidence %in% c("M","L"), confidence_note != "") |>
  select(Artist = name, Era = era, Region = region, Confidence = confidence,
         Composite = composite, `Confidence note` = confidence_note) |>
  arrange(Confidence, desc(Composite))

reactable(
  conf_notes,
  columns = list(
    Artist     = colDef(minWidth = 140, style = list(fontWeight = "600")),
    Era        = colDef(minWidth = 90),
    Region     = colDef(minWidth = 90),
    Confidence = colDef(minWidth = 80,
      style = function(value) {
        color <- if (value == "M") "#c8a000" else "#e08214"
        list(fontWeight = "700", color = color)
      }),
    Composite  = colDef(minWidth = 80),
    `Confidence note` = colDef(minWidth = 280)
  ),
  searchable = TRUE, striped = TRUE, highlight = TRUE, bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(background = "#2a1f2b", color = "#ffffff", fontWeight = "600")
  )
)
```

# 5 · Visualization Design

## Row {height="15%"}

::: {.callout-tip title="Teaching Prompt: The Same Data, Different Stories"}
**Every visualization choice is an argument.** The four charts below show the same data (composite scores by era) drawn four different ways. Each one highlights something and hides something else. Which is most honest? Which is most persuasive? Are those the same chart?
:::

## Row {height="85%"}

### A: Ranked bar {width="25%"}

```{r viz-bar}
era_summary <- artists |>
  group_by(era) |>
  summarise(mean_comp = round(mean(composite), 2), .groups = "drop")

p_bar <- era_summary |>
  ggplot(aes(x = fct_reorder(era, mean_comp), y = mean_comp, fill = era,
             text = paste0(era, ": ", mean_comp))) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = era_pal) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 10)) +
  labs(title = "A: Bar chart", subtitle = "Ranking emphasis", x = NULL, y = "Mean composite") +
  theme_hiphop()

ggplotly_titled(p_bar, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### B: Time trend line {width="25%"}

```{r viz-line}
p_line <- era_summary |>
  ggplot(aes(x = era, y = mean_comp, group = 1, color = era,
             text = paste0(era, ": ", mean_comp))) +
  geom_line(color = "#444444", linewidth = 1.1) +
  geom_point(size = 4, show.legend = FALSE) +
  scale_color_manual(values = era_pal) +
  scale_y_continuous(limits = c(5.5, 8.5)) +
  labs(title = "B: Line chart", subtitle = "Trend / change emphasis", x = NULL, y = "Mean composite") +
  theme_hiphop() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly_titled(p_line, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### C: Distribution (violin) {width="25%"}

```{r viz-violin}
era_levels <- levels(artists$era)

plot_ly(
  artists |> mutate(composite = as.numeric(composite)),
  x        = ~era,
  y        = ~composite,
  split    = ~era,
  type     = "violin",
  text     = ~paste0(name, ": ", composite),
  hoverinfo = "text",
  points   = "all",
  jitter   = 0.3,
  pointpos = 0,
  opacity  = 0.5,
  marker   = list(size = 4, opacity = 0.7),
  color    = ~era,
  colors   = era_pal,
  showlegend = FALSE
) |>
  layout(
    title  = list(text = "C: Violin + box · Distribution emphasis",
                  font = list(size = 13, color = "#3C3C3C")),
    xaxis  = list(title = "", categoryorder = "array",
                  categoryarray = era_levels),
    yaxis  = list(title = "Composite", range = c(1, 10.5)),
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)"
  )
```

### D: Scatter with n shown {width="25%"}

```{r viz-scatter-n}
era_n <- artists |> count(era, name = "n_artists")
era_summary2 <- left_join(era_summary, era_n, by = "era")

p_scatter_n <- era_summary2 |>
  ggplot(aes(x = era, y = mean_comp, size = n_artists, color = era,
             text = paste0(era, "\nMean: ", mean_comp, "\nN = ", n_artists, " artists"))) +
  geom_point(show.legend = FALSE, alpha = 0.85) +
  scale_color_manual(values = era_pal) +
  scale_size_continuous(range = c(4, 16)) +
  scale_y_continuous(limits = c(5.5, 8.5)) +
  labs(title = "D: Bubble chart", subtitle = "Shows sample size", x = NULL, y = "Mean composite") +
  theme_hiphop() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly_titled(p_scatter_n, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

# 6 · Careers & Longevity

## Row {height="15%"}

::: {.callout-warning title="Teaching Prompt: Survivorship & the Canon"}
**Does the composite measure quality, or career length?** Critical writing accumulates over a career: the longer an act stays active, the more analyses exist to score them with. Watch how scores drift upward with career length, then ask which way the causality runs. And notice how quickly most acts produce their signature work. If the defining album usually lands within a few years of debut, what exactly is a thirty-year career adding to the number? Age cuts the same question a different way: if the defining work reliably arrives in an artist's mid-twenties, era after era, what does that say about how the genre treats its elders?
:::

## Row {height="42.5%"}

### Career length vs. composite {width="55%"}

```{r career-scatter}
career_data <- artists |>
  mutate(
    status = if_else(is.na(active_to), "Active", "Concluded"),
    span_label = paste0(debut, "–", if_else(is.na(active_to), "present", as.character(active_to)))
  )

p_career <- career_data |>
  ggplot(aes(x = career_len, y = composite, color = era,
             text = paste0(
               "<b>", name, "</b><br>",
               "Active: ", span_label, " (", career_len, " yrs, ", status, ")<br>",
               "Signature: ", if_else(is.na(sig_work), "none recorded", sig_work), "<br>",
               "Composite: ", composite
             ))) +
  geom_jitter(width = 0.2, height = 0.05, alpha = 0.75, size = 2.5) +
  scale_color_manual(values = era_pal) +
  labs(
    title = "Career length vs. composite score",
    subtitle = "Length = debut through last active year (or present). Does longevity buy critical standing?",
    x = "Career length (years)", y = "Composite score", color = "Era"
  ) +
  theme_hiphop() +
  theme(legend.position = "right")

ggplotly_titled(p_career, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### Years from debut to signature work {width="45%"}

```{r years-to-peak}
peak_data   <- artists |> filter(!is.na(years_to_peak))
med_peak    <- median(peak_data$years_to_peak)
peak_counts <- peak_data |> count(years_to_peak)

p_peak <- peak_counts |>
  ggplot(aes(x = years_to_peak, y = n,
             text = paste0(years_to_peak, " yrs after debut: ", n, " acts"))) +
  geom_col(fill = "#6f9ec2", color = "white", linewidth = 0.3, width = 1) +
  geom_vline(xintercept = med_peak, linetype = "dashed", color = "#cf8079", linewidth = 0.8) +
  annotate("text", x = med_peak + 0.5, y = max(peak_counts$n) * 0.95, hjust = 0,
           label = paste0("median: ", med_peak, " yrs"), color = "#cf8079", size = 3.5) +
  labs(
    title = "How fast does the defining work arrive?",
    subtitle = "Years from debut to consensus signature release",
    x = "Years from debut to signature work", y = "Number of acts"
  ) +
  theme_hiphop()

ggplotly_titled(p_peak, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

## Row {height="42.5%"}

### Age at debut, by era {width="50%"}

```{r age-debut-era}
# Solo artists only: Compounds carry no Birth Year (members vary in age).
age_data <- artists |> filter(!is.na(birth))

p_age_debut <- age_data |>
  ggplot(aes(x = era, y = age_debut)) +
  geom_boxplot(aes(group = era), fill = "#d4b15a", alpha = 0.35,
               color = "#7a6a7b", outlier.shape = NA) +
  geom_jitter(aes(color = era,
                  text = paste0("<b>", name, "</b><br>",
                                "Born ", birth, ", debut ", debut,
                                " at age ", age_debut)),
              width = 0.18, height = 0.1, alpha = 0.75, size = 2) +
  scale_color_manual(values = era_pal, guide = "none") +
  labs(
    title = "Age at debut, by era",
    subtitle = "Is the genre debuting younger over time?",
    x = NULL, y = "Age at first commercial release"
  ) +
  theme_hiphop() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly_titled(p_age_debut, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### Age at signature work vs. composite {width="50%"}

```{r age-sig-composite}
sig_age_data <- age_data |> filter(!is.na(age_sig))
med_age_sig  <- median(sig_age_data$age_sig)

p_age_sig <- sig_age_data |>
  ggplot(aes(x = age_sig, y = composite, color = era,
             text = paste0("<b>", name, "</b><br>",
                           "Signature: ", sig_work, " at age ", age_sig, "<br>",
                           "Composite: ", composite))) +
  geom_jitter(width = 0.15, height = 0.05, alpha = 0.75, size = 2.5) +
  geom_vline(xintercept = med_age_sig, linetype = "dashed",
             color = "#7a6a7b", linewidth = 0.7) +
  scale_color_manual(values = era_pal) +
  scale_x_continuous(breaks = seq(15, 45, 5)) +
  labs(
    title = "Age at signature work vs. composite",
    subtitle = paste0("Do prodigies score higher? Median signature-work age: ",
                      med_age_sig),
    x = "Age when the signature work landed", y = "Composite score", color = "Era"
  ) +
  theme_hiphop() +
  theme(legend.position = "right")

ggplotly_titled(p_age_sig, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

# 7 · Full Data Table

## Row

```{r full-table}
display_data <- artists |>
  select(
    Symbol = sym, Artist = name, Region = region, Era = era, Style = style,
    Debut = debut, Rhyme = rhyme, Vocab = vocab, Story = story,
    Metaphor = metaphor, Concept = concept, Composite = composite,
    Conf = confidence
  ) |>
  arrange(Era, desc(Composite))

reactable(
  display_data,
  columns = list(
    Symbol    = colDef(maxWidth = 65, style = list(fontWeight = "700", color = "#6b4f6c")),
    Artist    = colDef(minWidth = 150, style = list(fontWeight = "600")),
    Region    = colDef(maxWidth = 110),
    Era       = colDef(maxWidth = 110),
    Style     = colDef(minWidth = 145),
    Debut     = colDef(maxWidth = 70),
    Rhyme     = colDef(maxWidth = 65,
      style = function(v) list(background = colorRampPalette(c("#fdf0ec","#b05a42"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Vocab     = colDef(maxWidth = 65,
      style = function(v) list(background = colorRampPalette(c("#f3edf3","#5d4360"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Story     = colDef(maxWidth = 65,
      style = function(v) list(background = colorRampPalette(c("#edf4ef","#3f6b50"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Metaphor  = colDef(maxWidth = 75,
      style = function(v) list(background = colorRampPalette(c("#ecf3f8","#33607f"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Concept   = colDef(maxWidth = 75,
      style = function(v) list(background = colorRampPalette(c("#f8f3e4","#7a5f22"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Composite = colDef(maxWidth = 85, style = list(fontWeight = "700", background = "#2a1f2b", color = "#fff")),
    Conf      = colDef(maxWidth = 55,
      style = function(v) {
        color <- switch(v, H = "#41ab5d", M = "#c8a000", L = "#e08214", "#000")
        list(fontWeight = "700", color = color)
      })
  ),
  searchable = TRUE, filterable = TRUE, sortable = TRUE,
  striped = TRUE, highlight = TRUE, bordered = TRUE,
  defaultPageSize = 25, showPageSizeOptions = TRUE,
  theme = reactableTheme(
    headerStyle = list(background = "#2a1f2b", color = "#ffffff", fontWeight = "600")
  )
)
```
