---
title: "Hip-Hop Periodic Table"
subtitle: "A Data Storytelling Teaching Module"
author: "Data Storytelling in Professional Practice"
format:
  dashboard:
    theme: 
      - cosmo
      - ../custom.scss
    logo: ""
    nav-buttons:
      - icon: github
        href: "https://github.com/lwylie01/my-site"
    scrolling: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(scales)
library(forcats)
library(stringr)
library(reactable)
library(htmltools)
library(readxl)

# ── Load data from Excel ───────────────────────────────────────────────────────
# File lives at: my-site/hiphop/data/hiphop_artists.xlsx
# To add new artists: open the Excel file, append rows to the "Artist Database"
# sheet, then push to main — the publish.yml GitHub Action rebuilds the site.

artists_raw <- read_excel(
  "data/hiphop_artists.xlsx",
  sheet = "Artist Database"
) |>
  # The dataset mixes individual artists ("Element") with groups ("Compound",
  # e.g. Wu-Tang Clan). This dashboard analyses individuals only; groups are
  # shown separately in the periodic table.
  filter(Type == "Element") |>
  rename(
    sym            = Symbol,
    name           = `Artist Name`,
    region         = Region,
    era            = Era,
    style          = Style,
    debut          = `Debut Year`,
    rhyme          = `Rhyme Density`,
    vocab          = `Vocab Breadth`,
    story          = Storytelling,
    metaphor       = `Metaphor/Imagery`,
    concept        = `Conceptual Depth`,
    confidence     = Confidence,
    confidence_note = `Confidence Note`,
    gender         = Gender
  ) |>
  mutate(across(c(rhyme, vocab, story, metaphor, concept), as.integer))

# ── Derived variables ─────────────────────────────────────────────────────────
artists <- artists_raw |>
  mutate(
    composite = round((rhyme + vocab + story + metaphor + concept) / 5, 1),
    era = factor(era, levels = c("Old School","Golden Age","Late 90s","2000s","2010s","2020s")),
    confidence = factor(confidence, levels = c("H","M","L")),
    region = factor(region, levels = c("NYC","Philly","LA","South","Midwest","Bay Area")),
    style = factor(style),
    confidence_note = if_else(is.na(confidence_note), "", confidence_note),
    strongest_dim = case_when(
      rhyme == pmax(rhyme, vocab, story, metaphor, concept) ~ "Rhyme Density",
      vocab == pmax(rhyme, vocab, story, metaphor, concept) ~ "Vocab Breadth",
      story == pmax(rhyme, vocab, story, metaphor, concept) ~ "Storytelling",
      metaphor == pmax(rhyme, vocab, story, metaphor, concept) ~ "Metaphor/Imagery",
      TRUE ~ "Conceptual Depth"
    )
  )

# ── Colour palettes ──────────────────────────────────────────────────────────
region_pal <- c(
  "NYC"      = "#3182bd",
  "Philly"   = "#4e9e94",
  "LA"       = "#e6550d",
  "South"    = "#31a354",
  "Midwest"  = "#756bb1",
  "Bay Area" = "#d0a400"
)

era_pal <- c(
  "Old School" = "#4292c6",
  "Golden Age" = "#41ab5d",
  "Late 90s"   = "#f6c141",
  "2000s"      = "#e08214",
  "2010s"      = "#9970ab",
  "2020s"      = "#d73027"
)

dim_pal <- c(
  "Rhyme Density"    = "#4472C4",
  "Vocab Breadth"    = "#70AD47",
  "Storytelling"     = "#ED7D31",
  "Metaphor/Imagery" = "#9B59B6",
  "Conceptual Depth" = "#C0392B"
)

conf_pal <- c("H" = "#41ab5d", "M" = "#f6c141", "L" = "#e08214")

# ── Shared ggplot2 theme ─────────────────────────────────────────────────────
theme_hiphop <- function() {
  theme_minimal(base_family = "sans", base_size = 12) +
    theme(
      plot.title    = element_text(size = 14, face = "bold", color = "#1a1a2e"),
      plot.subtitle = element_text(size = 11, color = "#555555"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#eeeeee"),
      axis.text     = element_text(color = "#444444"),
      legend.position = "bottom",
      legend.title  = element_text(size = 10, face = "bold"),
      strip.text    = element_text(face = "bold", color = "#333333")
    )
}
```

# Overview {.sidebar}

```{r sidebar-summary}
n_artists  <- nrow(artists)
n_high     <- sum(artists$confidence == "H")
n_medium   <- sum(artists$confidence == "M")
n_low      <- sum(artists$confidence == "L")
avg_comp   <- round(mean(artists$composite), 2)
top_artist <- artists |> slice_max(composite, n = 1) |> pull(name)

tagList(
  tags$div(style = "padding: 1rem 0;",
    tags$h4("Dataset at a Glance", style = "color:#1a1a2e; font-weight:700;"),
    tags$hr(style = "border-color:#dddddd;"),
    tags$p(tags$strong(n_artists), " artists across 6 eras"),
    tags$p(tags$strong(length(levels(artists$region))), " geographic regions"),
    tags$p(tags$strong(length(levels(artists$style))), " lyrical styles"),
    tags$hr(style = "border-color:#dddddd;"),
    tags$h5("Confidence breakdown", style = "color:#555;"),
    tags$p(
      tags$span(style = "color:#41ab5d; font-weight:700;", paste0("H: ", n_high, " artists")),
      tags$br(),
      tags$span(style = "color:#c8a000; font-weight:700;", paste0("M: ", n_medium, " artists (provisional)")),
      tags$br(),
      tags$span(style = "color:#e08214; font-weight:700;", paste0("L: ", n_low, " artists (estimates only)"))
    ),
    tags$hr(style = "border-color:#dddddd;"),
    tags$h5("Mean composite score", style = "color:#555;"),
    tags$p(tags$strong(avg_comp, " / 10")),
    tags$p(tags$em(paste0("Highest: ", top_artist))),
    tags$hr(style = "border-color:#dddddd;"),
    tags$h5("Teaching concepts", style = "color:#555; margin-top:1rem;"),
    tags$ul(style = "padding-left:1.2rem; font-size:0.85rem;",
      tags$li("Measurement validity"),
      tags$li("Sampling & bias"),
      tags$li("Multidimensional scoring"),
      tags$li("Composite vs. dimensions"),
      tags$li("Uncertainty communication"),
      tags$li("Visualization design")
    )
  )
)
```

# 1 · What Are We Measuring?

## Row {height="15%"}

::: {.callout-tip title="Teaching Prompt — Construct Validity"}
**The central measurement problem:** We started with one variable called "Lyricism" and ended up with five. Was that the right decision? What are we gaining and losing? This is the construct validity question that underlies every performance measure, survey instrument, and rating scale in your organization.
:::

## Row {height="85%"}

### Radar: Single artist — five dimensions {width="40%"}

```{r radar-single}
radar_artists <- c("Kendrick Lamar", "Scarface", "MF DOOM", "Rakim")

radar_data <- artists |>
  filter(name %in% radar_artists) |>
  select(name, rhyme, vocab, story, metaphor, concept) |>
  pivot_longer(-name, names_to = "dimension", values_to = "score") |>
  mutate(dimension = recode(dimension,
    rhyme    = "Rhyme Density",
    vocab    = "Vocab Breadth",
    story    = "Storytelling",
    metaphor = "Metaphor/Imagery",
    concept  = "Conceptual Depth"
  ))

fig_radar <- plot_ly(type = "scatterpolar", fill = "toself", mode = "lines+markers")

colors_radar <- c("#4472C4","#C0392B","#70AD47","#E08214")
for (i in seq_along(radar_artists)) {
  d <- radar_data |> filter(name == radar_artists[i])
  fig_radar <- fig_radar |> add_trace(
    r = c(d$score, d$score[1]),
    theta = c(d$dimension, d$dimension[1]),
    name = radar_artists[i],
    line = list(color = colors_radar[i], width = 2),
    fillcolor = paste0(colors_radar[i], "22")
  )
}

fig_radar |>
  layout(
    polar = list(radialaxis = list(visible = TRUE, range = c(0, 10), tickfont = list(size = 10))),
    legend = list(orientation = "h", x = 0.1, y = -0.1),
    margin = list(t = 10, b = 10),
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)"
  )
```

### Scatter: Composite score vs. Storytelling {width="60%"}

```{r scatter-composite-v-story}
p_scatter <- artists |>
  filter(confidence != "L") |>
  ggplot(aes(x = story, y = composite,
             color = era, size = composite,
             text = paste0(
               "<b>", name, "</b><br>",
               "Era: ", era, "<br>",
               "Composite: ", composite, "<br>",
               "Storytelling: ", story, "<br>",
               "Rhyme Density: ", rhyme
             ))) +
  geom_jitter(alpha = 0.75, width = 0.15, height = 0.05) +
  scale_color_manual(values = era_pal) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(2, 10.5), breaks = seq(2, 10, 2)) +
  labs(
    title = "Composite score vs. Storytelling dimension",
    subtitle = "Artists with the same composite can have very different dimension profiles",
    x = "Storytelling score (1–10)",
    y = "Composite score",
    color = "Era"
  ) +
  theme_hiphop() +
  theme(legend.position = "right")

ggplotly(p_scatter, tooltip = "text") |>
  layout(
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)"
  )
```

# 2 · Who Is In the Data?

## Row {height="15%"}

::: {.callout-warning title="Teaching Prompt — Sampling Decisions and Bias"}
**Every dataset has a frame — and a gap.** This dataset over-represents critically acclaimed artists, NYC, the Golden Age, and men. Those are documented choices. What datasets do you use at work where the frame is undocumented? What populations are missing?
:::

## Row {height="45%"}

### Artists by region {width="50%"}

```{r bar-region}
p_region <- artists |>
  count(region, era) |>
  ggplot(aes(x = fct_rev(region), y = n, fill = era,
             text = paste0(era, ": ", n, " artists"))) +
  geom_col(position = "stack") +
  scale_fill_manual(values = era_pal) +
  coord_flip() +
  labs(
    title = "Artists by region and era",
    subtitle = "NYC and the South dominate — international scenes sit outside the frame",
    x = NULL, y = "Number of artists", fill = "Era"
  ) +
  theme_hiphop()

ggplotly(p_region, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### Gender representation gap {width="50%"}

```{r bar-gender}
# Gender comes from the Gender column in the Excel workbook (single source of
# truth). "Mixed" only occurs for Compounds, which this dashboard filters out.
gender_data <- artists |>
  count(era, gender_group = gender)

n_female <- sum(artists$gender == "Female")

p_gender <- gender_data |>
  ggplot(aes(x = era, y = n, fill = gender_group,
             text = paste0(gender_group, ": ", n, " artists"))) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c("Female" = "#e377c2", "Male" = "#7f7f7f")) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Gender representation by era",
    subtitle = paste0(
      n_female, " female artists of ", nrow(artists), " total (",
      round(n_female/nrow(artists)*100), "%)"
    ),
    x = NULL, y = "Share of era", fill = NULL
  ) +
  theme_hiphop()

ggplotly(p_gender, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

## Row {height="40%"}

### Confidence distribution {width="35%"}

```{r confidence-donut}
conf_counts <- artists |>
  count(confidence) |>
  mutate(
    label = case_when(
      confidence == "H" ~ paste0("High\n", n, " artists"),
      confidence == "M" ~ paste0("Medium\n", n, " artists"),
      confidence == "L" ~ paste0("Low\n", n, " artists")
    )
  )

plot_ly(conf_counts,
  labels = ~label, values = ~n,
  type = "pie", hole = 0.55,
  marker = list(colors = c("#41ab5d","#f6c141","#e08214"),
                line = list(color = "#ffffff", width = 2)),
  textinfo = "label+percent",
  hoverinfo = "label+value"
) |>
  layout(
    title = list(text = "Scoring confidence distribution", font = list(size = 13)),
    showlegend = FALSE,
    paper_bgcolor = "rgba(0,0,0,0)"
  )
```

### What bias looks like in practice {width="65%"}

```{r bias-table}
bias_df <- tribble(
  ~"Bias type",        ~"Direction",                              ~"Practical implication",
  "Prestige",         "Over-represents critically acclaimed artists", "Commercial artists scored lower; vibe-first artists may be underrated",
  "Gender",           "~18% female vs. ~50% of rap audience",    "Female artists face different critical frameworks; scores may not be comparable",
  "Geography",        "NYC + Golden Age disproportionate",        "Southern and international artists underrepresented in critical canon used for scoring",
  "Language",         "English-language, US-based frame only",    "International and non-Anglophone rap is excluded from the sampling frame entirely — a documented scope decision, not an oversight",
  "Survivorship",     "Only artists who entered critical canon",  "Regionally significant artists without national coverage are absent",
  "Recency (inverted)","Post-2020 artists scored conservatively", "Emerging talent may be systematically underscored; revisit annually"
)

reactable(
  bias_df,
  columns = list(
    "Bias type"             = colDef(minWidth = 100, style = list(fontWeight = "600")),
    "Direction"             = colDef(minWidth = 160),
    "Practical implication" = colDef(minWidth = 200)
  ),
  striped = TRUE, highlight = TRUE, bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(background = "#1F3864", color = "#ffffff", fontWeight = "600")
  )
)
```

# 3 · Dimensions vs. Composite

## Row {height="15%"}

::: {.callout-note title="Teaching Prompt — The Aggregation Problem"}
**What does averaging destroy?** Rakim (Rhyme Density 10, composite 8.8) and Scarface (Storytelling 10, composite 8.2) land within half a point of each other on the composite — yet their dimension profiles look nothing alike. Is that near-equivalence meaningful, or does the composite erase the most important information? When does aggregation help, and when does it mislead?
:::

## Row {height="45%"}

### Heatmap: All artists, all dimensions {width="65%"}

```{r heatmap-dims}
dim_long <- artists |>
  filter(confidence == "H") |>
  arrange(era, composite) |>
  select(name, era, rhyme, vocab, story, metaphor, concept) |>
  pivot_longer(rhyme:concept, names_to = "dimension", values_to = "score") |>
  mutate(
    dimension = recode(dimension,
      rhyme    = "Rhyme",
      vocab    = "Vocab",
      story    = "Story",
      metaphor = "Metaphor",
      concept  = "Concept"
    ),
    dimension = factor(dimension, levels = c("Rhyme","Vocab","Story","Metaphor","Concept")),
    name = factor(name, levels = unique(name))
  )

p_heat <- dim_long |>
  ggplot(aes(x = dimension, y = name, fill = score,
             text = paste0(name, " — ", dimension, ": ", score))) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low = "#f7fbff", mid = "#6baed6", high = "#08306b",
    midpoint = 5.5, limits = c(1, 10),
    name = "Score"
  ) +
  facet_grid(era ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Dimension scores by artist (High-confidence only)",
    subtitle = "Each column is a dimension; each row an artist; color = score 1–10",
    x = NULL, y = NULL
  ) +
  theme_hiphop() +
  theme(
    axis.text.y = element_text(size = 7),
    strip.text.y = element_text(angle = 0, size = 8),
    panel.spacing = unit(0.3, "lines")
  )

ggplotly(p_heat, tooltip = "text", height = 900) |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### Era averages by dimension {width="35%"}

```{r era-dim-lines}
era_dim_avg <- artists |>
  group_by(era) |>
  summarise(
    `Rhyme Density`    = mean(rhyme),
    `Vocab Breadth`    = mean(vocab),
    `Storytelling`     = mean(story),
    `Metaphor/Imagery` = mean(metaphor),
    `Conceptual Depth` = mean(concept),
    .groups = "drop"
  ) |>
  pivot_longer(-era, names_to = "dimension", values_to = "avg")

p_era_lines <- era_dim_avg |>
  ggplot(aes(x = era, y = avg, color = dimension, group = dimension,
             text = paste0(dimension, " (", era, "): ", round(avg, 1)))) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = dim_pal) +
  scale_y_continuous(limits = c(4, 9), breaks = seq(4, 9, 1)) +
  labs(
    title = "Dimension averages across eras",
    subtitle = "Which dimensions rose or fell over time?",
    x = NULL, y = "Mean score", color = NULL
  ) +
  theme_hiphop() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly(p_era_lines, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

## Row {height="40%"}

### Strongest dimension by style {width="50%"}

```{r strongest-dim-style}
strongest <- artists |>
  count(style, strongest_dim) |>
  group_by(style) |>
  mutate(pct = n / sum(n))

p_strongest <- strongest |>
  ggplot(aes(x = style, y = pct, fill = strongest_dim,
             text = paste0(strongest_dim, ": ", round(pct*100), "%"))) +
  geom_col(position = "fill") +
  scale_fill_manual(values = dim_pal) +
  scale_y_continuous(labels = percent) +
  coord_flip() +
  labs(
    title = "Dominant dimension by style",
    subtitle = "What does each style specialize in?",
    x = NULL, y = "Share", fill = "Strongest dimension"
  ) +
  theme_hiphop() +
  theme(legend.position = "bottom")

ggplotly(p_strongest, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### Top 20 — composite score (High confidence) {width="50%"}

```{r top20-bar}
top20 <- artists |>
  filter(confidence == "H") |>
  slice_max(composite, n = 20) |>
  arrange(composite) |>
  mutate(name = factor(name, levels = name))

p_top20 <- top20 |>
  ggplot(aes(x = name, y = composite, fill = era,
             text = paste0(name, "\n", era, " | ", style, "\nComposite: ", composite))) +
  geom_col() +
  scale_fill_manual(values = era_pal) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 10.5)) +
  labs(
    title = "Top 20 composite scores (High confidence)",
    subtitle = "Composite = unweighted average of 5 dimensions",
    x = NULL, y = "Composite score (1–10)", fill = "Era"
  ) +
  theme_hiphop()

ggplotly(p_top20, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

# 4 · Uncertainty in Practice

## Row {height="15%"}

::: {.callout-important title="Teaching Prompt — Communicating Uncertainty"}
**Most dashboards show you the number. This one shows you how much to trust it.** The confidence flag is almost never present in stakeholder reporting — but it should be. Where in your organization's reporting does uncertainty get hidden? What decisions are being made on L-confidence data presented as H-confidence?
:::

## Row {height="45%"}

### Composite distribution by confidence tier {width="50%"}

```{r violin-confidence}
violin_data <- artists |>
  mutate(
    conf_label = factor(
      case_when(confidence == "H" ~ "High", confidence == "M" ~ "Medium", TRUE ~ "Low"),
      levels = c("High", "Medium", "Low")
    ),
    composite = as.numeric(composite)
  )

plot_ly(
  violin_data,
  x        = ~conf_label,
  y        = ~composite,
  split    = ~conf_label,
  type     = "violin",
  text     = ~paste0(name, ": ", composite),
  hoverinfo = "text",
  points   = "all",
  jitter   = 0.3,
  pointpos = 0,
  opacity  = 0.6,
  marker   = list(size = 5, opacity = 0.8),
  color    = ~conf_label,
  colors   = c("High" = "#41ab5d", "Medium" = "#f6c141", "Low" = "#e08214"),
  showlegend = FALSE
) |>
  layout(
    title  = list(text = "Score distribution by confidence tier",
                  font = list(size = 14, color = "#1a1a2e")),
    xaxis  = list(title = "Confidence tier"),
    yaxis  = list(title = "Composite score", range = c(1, 10.5)),
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)"
  )
```

### Score range: What we don't know {width="50%"}

```{r uncertainty-range}
uncertain <- artists |>
  filter(confidence %in% c("M","L")) |>
  arrange(desc(composite)) |>
  head(25) |>
  mutate(
    lo = pmax(1,  composite - if_else(confidence == "M", 0.8, 1.6)),
    hi = pmin(10, composite + if_else(confidence == "M", 0.8, 1.6)),
    name = fct_reorder(name, composite)
  )

p_range <- uncertain |>
  ggplot(aes(y = name, color = confidence,
             text = paste0(name, "\nScore: ", composite,
                           "\nPlausible range: ", round(lo,1), "–", round(hi,1)))) +
  geom_segment(aes(x = lo, xend = hi, yend = name), linewidth = 1.8, alpha = 0.4) +
  geom_point(aes(x = composite), size = 3) +
  scale_color_manual(values = conf_pal,
                     labels = c("M" = "Medium (±0.8)", "L" = "Low (±1.6)")) +
  scale_x_continuous(limits = c(1, 10), breaks = 1:10) +
  labs(
    title = "Plausible score ranges for provisional artists",
    subtitle = "Point = scored value; bar = plausible range given evidence quality",
    x = "Composite score", y = NULL, color = "Confidence tier"
  ) +
  theme_hiphop()

ggplotly(p_range, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

## Row {height="40%"}

### Confidence notes — M and L entries {width="100%"}

```{r conf-notes-table}
conf_notes <- artists |>
  filter(confidence %in% c("M","L"), confidence_note != "") |>
  select(Artist = name, Era = era, Region = region, Confidence = confidence,
         Composite = composite, `Confidence note` = confidence_note) |>
  arrange(Confidence, desc(Composite))

reactable(
  conf_notes,
  columns = list(
    Artist     = colDef(minWidth = 140, style = list(fontWeight = "600")),
    Era        = colDef(minWidth = 90),
    Region     = colDef(minWidth = 90),
    Confidence = colDef(minWidth = 80,
      style = function(value) {
        color <- if (value == "M") "#c8a000" else "#e08214"
        list(fontWeight = "700", color = color)
      }),
    Composite  = colDef(minWidth = 80),
    `Confidence note` = colDef(minWidth = 280)
  ),
  searchable = TRUE, striped = TRUE, highlight = TRUE, bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(background = "#1F3864", color = "#ffffff", fontWeight = "600")
  )
)
```

# 5 · Visualization Design

## Row {height="15%"}

::: {.callout-tip title="Teaching Prompt — The Same Data, Different Stories"}
**Every visualization choice is an argument.** The four charts below show the same underlying data — composite scores by era — rendered four different ways. Each emphasizes something different and hides something else. Which one is most honest? Which is most persuasive? Are those the same chart?
:::

## Row {height="85%"}

### A: Ranked bar {width="25%"}

```{r viz-bar}
era_summary <- artists |>
  group_by(era) |>
  summarise(mean_comp = round(mean(composite), 2), .groups = "drop")

p_bar <- era_summary |>
  ggplot(aes(x = fct_reorder(era, mean_comp), y = mean_comp, fill = era,
             text = paste0(era, ": ", mean_comp))) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = era_pal) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 10)) +
  labs(title = "A: Bar chart", subtitle = "Ranking emphasis", x = NULL, y = "Mean composite") +
  theme_hiphop()

ggplotly(p_bar, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### B: Time trend line {width="25%"}

```{r viz-line}
p_line <- era_summary |>
  ggplot(aes(x = era, y = mean_comp, group = 1, color = era,
             text = paste0(era, ": ", mean_comp))) +
  geom_line(color = "#444444", linewidth = 1.1) +
  geom_point(size = 4, show.legend = FALSE) +
  scale_color_manual(values = era_pal) +
  scale_y_continuous(limits = c(5.5, 8.5)) +
  labs(title = "B: Line chart", subtitle = "Trend / change emphasis", x = NULL, y = "Mean composite") +
  theme_hiphop() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly(p_line, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

### C: Distribution (violin) {width="25%"}

```{r viz-violin}
era_levels <- levels(artists$era)

plot_ly(
  artists |> mutate(composite = as.numeric(composite)),
  x        = ~era,
  y        = ~composite,
  split    = ~era,
  type     = "violin",
  text     = ~paste0(name, ": ", composite),
  hoverinfo = "text",
  points   = "all",
  jitter   = 0.3,
  pointpos = 0,
  opacity  = 0.5,
  marker   = list(size = 4, opacity = 0.7),
  color    = ~era,
  colors   = era_pal,
  showlegend = FALSE
) |>
  layout(
    title  = list(text = "C: Violin + box — Distribution emphasis",
                  font = list(size = 13, color = "#1a1a2e")),
    xaxis  = list(title = "", categoryorder = "array",
                  categoryarray = era_levels),
    yaxis  = list(title = "Composite", range = c(1, 10.5)),
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)"
  )
```

### D: Scatter with n shown {width="25%"}

```{r viz-scatter-n}
era_n <- artists |> count(era, name = "n_artists")
era_summary2 <- left_join(era_summary, era_n, by = "era")

p_scatter_n <- era_summary2 |>
  ggplot(aes(x = era, y = mean_comp, size = n_artists, color = era,
             text = paste0(era, "\nMean: ", mean_comp, "\nN = ", n_artists, " artists"))) +
  geom_point(show.legend = FALSE, alpha = 0.85) +
  scale_color_manual(values = era_pal) +
  scale_size_continuous(range = c(4, 16)) +
  scale_y_continuous(limits = c(5.5, 8.5)) +
  labs(title = "D: Bubble chart", subtitle = "Shows sample size", x = NULL, y = "Mean composite") +
  theme_hiphop() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly(p_scatter_n, tooltip = "text") |>
  layout(paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```

# 6 · Full Data Table

## Row

```{r full-table}
display_data <- artists |>
  select(
    Symbol = sym, Artist = name, Region = region, Era = era, Style = style,
    Debut = debut, Rhyme = rhyme, Vocab = vocab, Story = story,
    Metaphor = metaphor, Concept = concept, Composite = composite,
    Conf = confidence
  ) |>
  arrange(Era, desc(Composite))

reactable(
  display_data,
  columns = list(
    Symbol    = colDef(maxWidth = 65, style = list(fontWeight = "700", color = "#1F3864")),
    Artist    = colDef(minWidth = 150, style = list(fontWeight = "600")),
    Region    = colDef(maxWidth = 110),
    Era       = colDef(maxWidth = 110),
    Style     = colDef(minWidth = 145),
    Debut     = colDef(maxWidth = 70),
    Rhyme     = colDef(maxWidth = 65,
      style = function(v) list(background = colorRampPalette(c("#deebf7","#08519c"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Vocab     = colDef(maxWidth = 65,
      style = function(v) list(background = colorRampPalette(c("#e5f5e0","#006d2c"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Story     = colDef(maxWidth = 65,
      style = function(v) list(background = colorRampPalette(c("#feedde","#a63603"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Metaphor  = colDef(maxWidth = 75,
      style = function(v) list(background = colorRampPalette(c("#f2e2f7","#6a1a8a"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Concept   = colDef(maxWidth = 75,
      style = function(v) list(background = colorRampPalette(c("#fee0d2","#a50f15"))(10)[max(1,v)],
                                color = if(v >= 7) "#fff" else "#000")),
    Composite = colDef(maxWidth = 85, style = list(fontWeight = "700", background = "#1F3864", color = "#fff")),
    Conf      = colDef(maxWidth = 55,
      style = function(v) {
        color <- switch(v, H = "#41ab5d", M = "#c8a000", L = "#e08214", "#000")
        list(fontWeight = "700", color = color)
      })
  ),
  searchable = TRUE, filterable = TRUE, sortable = TRUE,
  striped = TRUE, highlight = TRUE, bordered = TRUE,
  defaultPageSize = 25, showPageSizeOptions = TRUE,
  theme = reactableTheme(
    headerStyle = list(background = "#1F3864", color = "#ffffff", fontWeight = "600")
  )
)
```
